% ══════════════════════════════════════════════════════════════════════════
%  Chapter 6: Linear Threshold Functions, Reed--Muller Codes, and
%             the Group-Theoretic Perspective
% ══════════════════════════════════════════════════════════════════════════

\chapter{Linear Threshold Functions, Codes, and Groups}\label{ch:ltf-codes-groups}

This chapter covers three topics that connect the Boolean Fourier theory
to the thesis: linear threshold functions (the basic computational unit
of binary networks), Reed--Muller codes (the coding-theoretic
interpretation of spectral truncation), and the group-theoretic
viewpoint (which places the Walsh--Fourier basis in a broader context).

% ══════════════════════════════════════════════════════════════════════
\section{Linear Threshold Functions}\label{sec:ltf}

\begin{definition}[Linear threshold function (LTF)]\label{def:ltf}
A \emph{linear threshold function} is a Boolean function of the form
\begin{equation}\label{eq:ltf}
  f(x) = \sgn\!\left(\sum_{i=1}^n w_i x_i\right)
  = \sgn(w^\top x),
\end{equation}
where $w = (w_1, \ldots, w_n) \in \R^n$ is a weight vector and
$\sgn(z) = +1$ if $z \geq 0$ and $-1$ if $z < 0$.
\end{definition}

LTFs are also called \emph{halfspaces} or \emph{threshold gates}.
In the neural network context, a single neuron with binary inputs and a
sign activation computes exactly an LTF. A binary neural network layer
computes $d$ LTFs in parallel (one per row of the weight matrix).

\begin{example}\label{ex:ltf-examples}
\begin{itemize}
  \item \textbf{Dictator:} $f(x) = \sgn(x_1) = x_1$. Weight vector
    $w = (1, 0, \ldots, 0)$.
  \item \textbf{Majority:} $f(x) = \sgn(x_1 + \cdots + x_n)$. Weight
    vector $w = (1, 1, \ldots, 1)$.
  \item \textbf{Weighted threshold:} $f(x) = \sgn(3x_1 + x_2 - 2x_3)$.
    This function outputs $+1$ when $3x_1 + x_2 - 2x_3 > 0$.
\end{itemize}
The parity function $x_1 x_2 \cdots x_n$ is \emph{not} an LTF for
$n \geq 2$ (it cannot be computed by a single threshold gate).
\end{example}


% ──────────────────────────────────────────────────────────────────────
\subsection{Chow's Theorem}\label{sec:chow}

\begin{theorem}[Chow, 1961; {\citealp{chow1961recognition}}; see also {\citet[Chapter~5]{odonnell2014analysis}}]
\label{thm:chow}
A linear threshold function $f$ is uniquely determined among all LTFs
by its degree-0 and degree-1 Fourier coefficients:
$\fhat(\emptyset) = \E[f]$ and $\fhat(\{i\}) = \E[f \cdot x_i]$ for
$i = 1, \ldots, n$.
\end{theorem}

The $n+1$ values $(\fhat(\emptyset), \fhat(\{1\}), \ldots, \fhat(\{n\}))$
are called the \emph{Chow parameters}. Chow's theorem says: if two LTFs
have the same Chow parameters, they are the same function.

\begin{remark}\label{rem:chow-significance}
Chow's theorem implies that the ``essential information'' of an LTF lives
in degree $\leq 1$ of the Fourier spectrum. This does not mean the
higher-degree coefficients are zero---they are generally nonzero---but
that they are \emph{determined} by the low-degree ones (within the class
of LTFs). This is a strong structural result supporting the idea that
LTFs are ``low-complexity'' functions.
\end{remark}

\begin{proof}[Proof sketch]
Two distinct LTFs $f = \sgn(w^\top x)$ and $g = \sgn(v^\top x)$ define
different halfspaces in $\R^n$. If they had the same Chow parameters,
then $\E[f \cdot x_i] = \E[g \cdot x_i]$ for all $i$, meaning $f$ and
$g$ agree on average with every coordinate projection.

The key geometric insight: the Chow parameters determine the orthogonal
projection of $f$ onto the degree-$1$ subspace of $L^2(\pmone^n)$. For
LTFs, this projection is essentially the (normalised) weight vector---one
can show $\fhat(\{i\}) = c \cdot w_i / \norm{w}_2$ for a constant $c$
depending only on $\norm{w}$---so two LTFs with the same Chow parameters
must have proportional weight vectors, hence define the same halfspace.
See \citet[Chapter~5]{odonnell2014analysis} for the full argument.
\end{proof}


% ──────────────────────────────────────────────────────────────────────
\subsection{Spectral Properties of LTFs}\label{sec:ltf-spectrum}

LTFs have several spectral properties that are relevant to the thesis:

\paragraph{Spectral concentration.}
LTFs are spectrally concentrated at low degrees. While the exact spectrum
depends on the weight vector $w$, the majority function (equal weights)
has 75\% of its energy at degree~1 for $n = 3$, and the concentration
improves with $n$. The total influence bound $\mathbf{I}[f] = O(\sqrt{n})$
(\Cref{thm:ltf-noise} below), combined with the level-$k$ inequality
(\Cref{thm:level-k}), gives the tail bound
$W^k[f] \leq (e\sqrt{n}/k)^k$, which becomes small when
$k \gg e\sqrt{n}$. So most of the spectral energy is within the first
$O(\sqrt{n})$ degrees.

\paragraph{Noise sensitivity.}
LTFs have bounded noise sensitivity. The following result quantifies this:

\begin{theorem}[{\citet{harsha2014bounding}}]\label{thm:ltf-noise}
For any LTF $f \colon \pmone^n \to \pmone$, the total influence satisfies
$\mathbf{I}[f] = O(\sqrt{n})$.
\end{theorem}

More precisely, for $f = \sgn(w^\top x)$ with
$\norm{w}_2 = 1$: $\mathbf{I}[f] = \Theta(\norm{w}_1)$ as
$n \to \infty$ (the asymptotic is in $n$, with the implicit constants
being absolute). The idea: each coordinate's influence equals the
probability that $x_i$ is ``pivotal'' (flipping $x_i$ changes the sign
of $w^\top x$), which can be estimated by the CLT applied to the sum
$w^\top x$ minus the $i$-th term. Berry--Ess\'een
(\Cref{thm:berry-esseen}) makes this rigorous. For binary weights
$w \in \pmone^n$, $\norm{w}_1 = n$ and $\norm{w}_2 = \sqrt{n}$, giving
$\mathbf{I}[f] \approx \sqrt{2n/\pi}$.

\paragraph{Relevance to the thesis.}
Each row of a binary weight matrix followed by a sign activation computes
an LTF. Note that here $n$ denotes the number of inputs to the LTF
(i.e., the layer width $d$ in the thesis, not $\log_2 d$). The bounded
noise sensitivity of LTFs means that:
\begin{itemize}
  \item Each neuron in a binary layer with $d$ inputs has total influence
    $O(\sqrt{d})$; at $d = 1024$ this is $\approx 25$.
  \item The spectral tail (high-degree coefficients) is bounded, supporting
    the spectral concentration hypothesis.
  \item The sign activation acts as a ``spectral filter,'' preserving
    low-degree structure and discarding high-frequency noise.
\end{itemize}

\begin{remark}[Variable naming]\label{rem:ltf-variable-naming}
Caution: in the LTF literature, $n$ denotes the number of input bits to
the function. In the thesis's spectral parameterization, each row of
a $d \times d$ weight matrix is viewed as a Boolean function on
$n = \log_2 d$ variables (the ``Fourier domain'' variables). These are
different uses of $n$. When we say an LTF on $d$ inputs has total
influence $O(\sqrt{d})$, we mean the LTF viewed as a function on $d$
literal input coordinates, \emph{not} on the $\log_2 d$ Fourier variables.
\end{remark}


% ══════════════════════════════════════════════════════════════════════
\section{Reed--Muller Codes}\label{sec:reed-muller}

There is a deep connection between the Boolean Fourier basis and
Reed--Muller codes from coding theory. The thesis uses this connection
to analyze the spectral-to-binary projection.

References: \citet{reed1954class,muller1954application} for the
original constructions; \citet{macwilliams1977theory} for a comprehensive
treatment.

\subsection{Definition}\label{sec:rm-def}

\begin{definition}[Reed--Muller code]\label{def:reed-muller}
Here $\text{GF}(2) = \{0, 1\}$ is the \emph{field with two elements},
where addition is XOR ($0 \oplus 1 = 1$, $1 \oplus 1 = 0$) and
multiplication is AND ($1 \cdot 1 = 1$, all others $0$). A
\emph{linear code} of length $N$ over $\text{GF}(2)$ is a linear
subspace of $\text{GF}(2)^N$; its \emph{dimension} is the dimension of
this subspace, and its \emph{minimum distance} is the minimum
\emph{Hamming distance} between distinct codewords (the Hamming
distance between two vectors is the number of coordinates in which
they differ).

The $r$-th order \emph{Reed--Muller code} $\text{RM}(r, n)$ consists of
all evaluation vectors of polynomials $p(b_1, \ldots, b_n)$ over
$\text{GF}(2)$ of total degree $\leq r$, evaluated at all $2^n$ points
of $\{0,1\}^n$.
\end{definition}

\paragraph{$\text{GF}(2)$ perspective.}
$\text{RM}(r, n)$ is a linear code of length $2^n$ over $\text{GF}(2)$.
Its dimension is $\sum_{i=0}^r \binom{n}{i}$ and its minimum distance
is $2^{n-r}$.

\paragraph{$\pmone$ perspective.}
We can convert to the $\pmone$ convention via $b \mapsto (-1)^b$. Under
this map, a $\text{GF}(2)$-monomial $b_{i_1} \cdots b_{i_k}$ (which
computes AND over $\{0,1\}$) does \emph{not} become the Walsh monomial
$\chi_{\{i_1, \ldots, i_k\}}(x) = x_{i_1} \cdots x_{i_k}$. Instead,
$b_i = \frac{1 - x_i}{2}$, so $\text{GF}(2)$-monomials of degree $k$
become $\pmone$-polynomials involving Walsh monomials of degree
$\leq k$ (but not only degree $k$).

Nevertheless, the \emph{span} is preserved: the vector space of
$\text{GF}(2)$-polynomials of degree $\leq r$, after the $b \mapsto (-1)^b$
map, coincides with the vector space spanned by $\{\chi_S : |S| \leq r\}$.
This is because both are $(r+1)$-level filtrations of the same
$2^n$-dimensional space, and the change-of-basis between the two is
\emph{unitriangular}: when expressing $\text{GF}(2)$-monomials in terms of
Walsh monomials (ordered by degree), the matrix is lower-triangular with
$1$s on the diagonal. Concretely, $b_i = (1 - x_i)/2$, so a
$\text{GF}(2)$-monomial of degree $k$ equals $\prod_{i \in S}(1-x_i)/2^k$,
which is $(\pm 1/2^k)\,\chi_S$ plus lower-degree Walsh terms. The
leading coefficient is nonzero, ensuring invertibility at each level.

Thus, $\text{RM}(r,n)$ in the $\pmone$ convention consists exactly of
those $\pmone$-vectors whose Walsh--Fourier expansion has degree $\leq r$.

\begin{theorem}[RM parameters]\label{thm:rm-params}
$\text{RM}(r, n)$ is a linear code with:
\begin{itemize}
  \item \emph{Length:} $N = 2^n$.
  \item \emph{Dimension:} $\sum_{i=0}^r \binom{n}{i}$.
  \item \emph{Minimum distance:} $2^{n-r}$.
\end{itemize}
\end{theorem}

\begin{example}\label{ex:rm-small}
\begin{itemize}
  \item $\text{RM}(0, n)$: repetition code. Only two codewords: all-$1$
    and all-$(-1)$. Minimum distance $2^n$. (Degree~0 functions are
    constants $\pm 1$.)
  \item $\text{RM}(1, n)$: the first-order RM code. Dimension $n + 1$.
    Codewords are the $\pmone$-evaluation vectors of affine functions
    $c_\emptyset \oplus c_1 b_1 \oplus \cdots \oplus c_n b_n$ over
    $\text{GF}(2)$, i.e., functions of Walsh--Fourier degree $\leq 1$.
    There are $2^{n+1}$ codewords. Minimum distance $2^{n-1}$.

    \textbf{Worked example: $\text{RM}(1, 2)$.}
    The affine functions over $\text{GF}(2)$ in two variables are:
    $0$, $1$, $b_1$, $1 \oplus b_1$, $b_2$, $1 \oplus b_2$,
    $b_1 \oplus b_2$, $1 \oplus b_1 \oplus b_2$---eight functions in
    all ($2^{n+1} = 2^3 = 8$). Their evaluation vectors over
    $\{0,1\}^2$, mapped to $\pmone$ via $0 \mapsto +1$, $1 \mapsto -1$:
    \begin{center}
    \begin{tabular}{lcc}
    \toprule
    Function & $\{0,1\}^2$ eval & $\pmone$ codeword \\
    \midrule
    $0$               & $(0,0,0,0)$ & $(+1,+1,+1,+1)$ \\
    $b_1$             & $(0,1,0,1)$ & $(+1,-1,+1,-1)$ \\
    $b_2$             & $(0,0,1,1)$ & $(+1,+1,-1,-1)$ \\
    $b_1 \oplus b_2$  & $(0,1,1,0)$ & $(+1,-1,-1,+1)$ \\
    \bottomrule
    \end{tabular}
    \end{center}
    (plus their bitwise complements for the remaining four).
    These are exactly the evaluation vectors of the degree-$\leq 1$
    Walsh functions: $1$, $x_1$, $x_2$, $x_1 x_2$ would be degree~2,
    so it is excluded. Minimum distance = $2^{2-1} = 2$.
  \item $\text{RM}(n, n)$: the full code, containing all $2^{2^n}$
    Boolean functions. Minimum distance~1.
\end{itemize}
\end{example}


% ──────────────────────────────────────────────────────────────────────
\subsection{Connection to the Thesis}\label{sec:rm-thesis}

The thesis's degree-$k$ spectral parameterization optimizes within the
span of $\{\chi_S : |S| \leq k\}$---which is exactly the ``continuous
relaxation'' of $\text{RM}(k, n)$. The coefficients are real during
training and are projected to $\pmone$ via the sign function.

Three properties of RM codes are relevant:

\begin{enumerate}[label=(\roman*)]
  \item \textbf{Minimum distance.}
    $\text{RM}(k, n)$ has minimum distance $2^{n-k}$. For the thesis
    with $k = 3$ and $n = 10$: minimum distance $= 2^7 = 128$. This
    means any two distinct degree-$\leq 3$ Boolean functions differ on
    at least 128 of the 1024 inputs.

  \item \textbf{Dimension = parameter count.}
    The dimension $\sum_{i=0}^k \binom{n}{i}$ is exactly the number of
    spectral coefficients in the thesis's parameterization:
    $\binom{10}{\leq 3} = 1 + 10 + 45 + 120 = 176$.

  \item \textbf{Sign projection is not RM decoding.}
    The sign function applied to a continuous approximation
    $f^{\leq k}(x)$ is \emph{not} guaranteed to produce a codeword of
    $\text{RM}(k, n)$. The sign of a degree-$k$ polynomial can have
    Fourier support up to degree~$n$. For example, the degree-1
    polynomial $p(x) = \frac{1}{2}(x_1 + x_2 + x_3)$ has
    $\sgn(p) = \text{Maj}_3$, which has a nonzero degree-3 Fourier
    coefficient $\widehat{\text{Maj}}_3(\{1,2,3\}) = -\frac{1}{2}$.
    The sign operation ``creates'' high-degree spectral content. The RM
    connection provides structural insight but does not directly
    constrain the output.
\end{enumerate}


% ══════════════════════════════════════════════════════════════════════
\section{The Group-Theoretic Perspective}\label{sec:group-theory}

The Boolean Fourier expansion is a special case of Fourier analysis on
finite abelian groups. This section briefly develops that perspective.
The main reference is \citet{terras1999fourier}; see also
\citet{diaconis1988group}.

\subsection{The Group $\Z_2^n$}\label{sec:z2n}

The Boolean hypercube $\pmone^n$ under coordinatewise multiplication is
isomorphic to the group $\Z_2^n = (\Z/2\Z)^n$ under coordinatewise
addition mod~2.

The isomorphism: $\phi \colon \Z_2^n \to \pmone^n$ by
$\phi(b_1, \ldots, b_n) = ((-1)^{b_1}, \ldots, (-1)^{b_n})$.
Then $\phi(a + b) = \phi(a) \cdot \phi(b)$ (coordinatewise).

\subsection{Characters of Finite Abelian Groups}\label{sec:characters}

\begin{definition}[Character]\label{def:character}
A \emph{character} of a finite abelian group $G$ is a group homomorphism
$\chi \colon G \to \mathbb{C}^*$ (the nonzero complex numbers under
multiplication). In general, characters are complex-valued, but for
$\Z_2^n$ they turn out to be real---in fact $\pmone$-valued---as the
proof below shows.
\end{definition}

\begin{theorem}\label{thm:characters-z2n}
The characters of $\Z_2^n$ are exactly the functions
$\chi_S(b) = (-1)^{\sum_{i \in S} b_i}$ for $S \subseteq [n]$.
Under the $\pmone$ convention, these become $\chi_S(x) = \prod_{i \in S} x_i$:
exactly the parity functions.
\end{theorem}

\begin{proof}
A character $\chi$ of $\Z_2^n$ is determined by its values on the
standard generators $e_i = (0, \ldots, 1, \ldots, 0)$. Since
$2 e_i = 0$ in $\Z_2^n$, $\chi(e_i)^2 = \chi(0) = 1$, so
$\chi(e_i) \in \{+1, -1\}$. The $2^n$ choices for
$(\chi(e_1), \ldots, \chi(e_n)) \in \pmone^n$ give all $2^n$ characters.
\end{proof}

\begin{theorem}[Fourier analysis on finite abelian groups]\label{thm:fourier-group}
For any finite abelian group $G$ with character group $\hat{G}$, the
characters form an orthonormal basis for $L^2(G)$ (functions $G \to \mathbb{C}$
with the inner product $\ip{f}{g} = \frac{1}{|G|}\sum_{x \in G} f(x)\overline{g(x)}$).
Every function $f \colon G \to \mathbb{C}$ has a unique expansion
$f = \sum_{\chi \in \hat{G}} \fhat(\chi)\, \chi$, and Parseval holds:
$\norm{f}_2^2 = \sum_\chi |\fhat(\chi)|^2$.
\end{theorem}

For $G = \Z_2^n$, this recovers exactly the Boolean Fourier expansion of
\Cref{ch:boolean-fourier}.

\subsection{Comparison Table}\label{sec:comparison-table}

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Classical ($\R/\Z$ or $\Z/N\Z$)} & \textbf{Boolean ($\Z_2^n$)} & \textbf{Role} \\
\midrule
$e^{2\pi i k x}$ & $\chi_S(x) = \prod_{i \in S} x_i$ & Basis function \\
Complex-valued & $\pm 1$-valued & Basis alphabet \\
DFT / FFT & WHT / FWHT & Fast algorithm \\
$O(N \log N)$ complex ops & $O(N \log N)$ integer add/sub & Complexity \\
$\sum_k |\hat{f}(k)|^2 = \norm{f}_2^2$ & $\sum_S \fhat(S)^2 = \norm{f}_2^2$ & Parseval \\
$\widehat{f \ast g}(k) = \hat{f}(k)\hat{g}(k)$ & $\widehat{f \ast g}(S) = \fhat(S)\ghat(S)$ & Convolution \\
Frequency $k$ & Degree $|S|$ & ``Frequency'' \\
Low-pass filter & Degree truncation & Smoothing \\
\bottomrule
\end{tabular}
\end{center}

The Boolean case is simpler in every respect: the basis is real (actually
$\pm 1$-valued), the transform uses only additions and subtractions, and
the group structure is particularly transparent ($\Z_2^n$ is a product of
copies of the simplest nontrivial group). This simplicity is precisely
what makes it attractive for binary neural networks.

\medskip
\paragraph{Connecting the three perspectives.}
All three topics in this chapter illuminate the same central phenomenon:
the Walsh--Fourier basis is the \emph{natural} tool for analyzing
binary-valued functions. LTFs show that the most basic computational
unit (a threshold gate) has low-degree spectral concentration. Reed--Muller
codes show that degree-bounded Walsh expansions form a classical
error-correcting code. And the group-theoretic viewpoint explains
\emph{why} the Walsh basis is canonical: it consists of exactly the
characters of the underlying group $\Z_2^n$, the unique orthonormal
basis that simultaneously diagonalizes all translation operators.


% ──────────────────────────────────────────────────────────────────────
\section*{Recommended Reading}

\begin{itemize}
  \item \textbf{\citet[Chapter~5]{odonnell2014analysis}.}
    Full proof of Chow's theorem and a detailed treatment of LTF
    spectral properties (influence bounds, noise sensitivity, the
    Gotsman--Linial conjecture). Graduate level.

  \item \textbf{\citet{macwilliams1977theory}, \emph{The Theory of
    Error-Correcting Codes}.}
    The encyclopedic reference for coding theory. Chapter~13 covers
    Reed--Muller codes in detail, including decoding algorithms and
    weight distributions. Graduate level; assumes abstract algebra.

  \item \textbf{\citet{terras1999fourier}, \emph{Fourier Analysis on
    Finite Groups and Applications}.}
    Develops Fourier analysis on arbitrary finite groups, with $\Z_2^n$
    as a running example. Chapters~1--4 provide the group-theoretic
    foundation for the character theory in \Cref{sec:characters}.
    Assumes undergraduate algebra (groups, homomorphisms).

  \item \textbf{\citet{diaconis1988group}, \emph{Group Representations
    in Probability and Statistics}.}
    A beautifully written monograph connecting group representations to
    random walks, card shuffling, and statistical applications. Lighter
    on Boolean analysis but excellent for building intuition about why
    Fourier analysis on groups works. Assumes basic group theory.
\end{itemize}
