\documentclass[12pt,a4paper]{report}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage{setspace}
\usepackage{natbib}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{makeidx}

\makeindex

% ── Cover colors ─────────────────────────────────────────────────────
\definecolor{covblue}{RGB}{41, 98, 255}
\definecolor{covteal}{RGB}{0, 150, 136}
\definecolor{covpurp}{RGB}{142, 36, 170}
\definecolor{covamber}{RGB}{245, 166, 35}

\onehalfspacing
\pagestyle{fancy}
\fancyhf{}
\fancyhead[R]{\textit{\small Boolean Fourier Analysis: An Expository Guide}}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% ── Theorem environments ──────────────────────────────────────────────
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% ── Macros ────────────────────────────────────────────────────────────
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\pmone}{\{-1,+1\}}
\newcommand{\fhat}{\widehat{f}}
\newcommand{\ghat}{\widehat{g}}
\newcommand{\ip}[2]{\langle #1, #2 \rangle}
\newcommand{\norm}[1]{\left\| #1 \right\|}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Inf}{\operatorname{Inf}}
\newcommand{\Stab}{\operatorname{Stab}}
\newcommand{\sgn}{\operatorname{sgn}}
\DeclareMathOperator{\WHT}{WHT}
\DeclareMathOperator{\FWHT}{FWHT}

\begin{document}

% ── Cover page ───────────────────────────────────────────────────────
\begin{titlepage}
\thispagestyle{empty}
\centering

\vspace*{0.8cm}

{\color{covblue}\rule{\linewidth}{2pt}}\\[10pt]
{\fontsize{28}{34}\selectfont\bfseries Boolean Fourier Analysis\\[6pt]
and the Walsh--Hadamard Transform}\\[10pt]
{\color{covblue}\rule{\linewidth}{0.6pt}}\\[6pt]
{\Large\bfseries An Expository Guide}\\[6pt]
{\color{covblue}\rule{\linewidth}{2pt}}

\vspace{0.9cm}

% ── Tesseract: the 4-dimensional Boolean hypercube {±1}^4 ────────
% Projection: (v1,v2,v3,v4) → 1.8·(v1 + φv3, v2 + φv4) where φ ≈ 0.618
\begin{tikzpicture}[scale=1]
  \pgfmathsetmacro{\phi}{0.618}
  \pgfmathsetmacro{\sc}{1.8}

  % Place all 16 vertices
  \foreach \i in {0,...,15} {
    \pgfmathsetmacro{\va}{1-2*Mod(\i,2)}
    \pgfmathsetmacro{\vb}{1-2*Mod(int(\i/2),2)}
    \pgfmathsetmacro{\vc}{1-2*Mod(int(\i/4),2)}
    \pgfmathsetmacro{\vd}{1-2*Mod(int(\i/8),2)}
    \pgfmathsetmacro{\px}{\sc*(\va + \phi*\vc)}
    \pgfmathsetmacro{\py}{\sc*(\vb + \phi*\vd)}
    \coordinate (V\i) at (\px, \py);
  }

  % Edges by dimension (8 edges each, 32 total)
  % Dimension 1 (bit 0): i even → edge i—(i+1)
  \foreach \i in {0,2,4,6,8,10,12,14} {
    \pgfmathtruncatemacro{\j}{\i+1}
    \draw[covblue, line width=1.3pt, opacity=0.8] (V\i) -- (V\j);
  }
  % Dimension 2 (bit 1): i mod 4 < 2 → edge i—(i+2)
  \foreach \i in {0,1,4,5,8,9,12,13} {
    \pgfmathtruncatemacro{\j}{\i+2}
    \draw[covteal, line width=1.3pt, opacity=0.8] (V\i) -- (V\j);
  }
  % Dimension 3 (bit 2): i mod 8 < 4 → edge i—(i+4)
  \foreach \i in {0,1,2,3,8,9,10,11} {
    \pgfmathtruncatemacro{\j}{\i+4}
    \draw[covpurp, line width=1.3pt, opacity=0.8] (V\i) -- (V\j);
  }
  % Dimension 4 (bit 3): i < 8 → edge i—(i+8)
  \foreach \i in {0,...,7} {
    \pgfmathtruncatemacro{\j}{\i+8}
    \draw[covamber, line width=1.3pt, opacity=0.8] (V\i) -- (V\j);
  }

  % Vertices
  \foreach \i in {0,...,15} {
    \fill[white] (V\i) circle (3.5pt);
    \fill[black!75] (V\i) circle (2.2pt);
  }

  % Label two antipodal vertices
  \node[above right=2pt, font=\footnotesize\itshape, black!70]
    at (V0) {$(1,1,1,1)$};
  \node[below left=2pt, font=\footnotesize\itshape, black!70]
    at (V15) {$(-1,-1,-1,-1)$};

  % Caption
  \node[below=8pt, font=\small] at (0,-3.3)
    {The Boolean hypercube $\{-1,+1\}^4$: 16 vertices, 32 edges, four dimensions};

  % Small legend
  \node[font=\scriptsize, right] at (2.2, -4.0)
    {{\color{covblue}---}\,$x_1$\quad
     {\color{covteal}---}\,$x_2$\quad
     {\color{covpurp}---}\,$x_3$\quad
     {\color{covamber}---}\,$x_4$};
\end{tikzpicture}

\vspace{0.8cm}

{\large\itshape From First Principles to Spectral Methods\\[3pt]
for Binary Neural Networks}

\vfill

{\large Companion to the thesis:\\[4pt]
\textit{Spectral Methods for Binary Neural Networks}}

\vspace{0.8cm}

{\large March 2026}

\vspace{0.5cm}

\end{titlepage}


% ── Dedication ───────────────────────────────────────────────────────
\newpage
\thispagestyle{empty}
\vspace*{\fill}
\begin{center}
\large\itshape
Dedicated to the $2^n$ inhabitants of $\{-1,+1\}^n$.
\end{center}
\vspace*{\fill}


% ── Epigraph ─────────────────────────────────────────────────────────
\newpage
\thispagestyle{empty}
\vspace*{3cm}

\begin{quote}
\large\itshape
``Mathematics compares the most diverse phenomena and discovers the
secret analogies that unite them.''
\end{quote}
\hfill --- Joseph Fourier, \textit{Th\'eorie analytique de la chaleur} (1822)

\vspace{2cm}

\begin{quote}
\large\itshape
``In mathematics you don't understand things. You just get used to them.''
\end{quote}
\hfill --- John von Neumann

\vspace{2cm}

\begin{quote}
\large\itshape
``The purpose of computing is insight, not numbers.''
\end{quote}
\hfill --- Richard Hamming, \textit{Numerical Methods for Scientists}\\
\hfill\textit{and Engineers} (1962)

\newpage


% ── Front matter ─────────────────────────────────────────────────────
\tableofcontents
\newpage

% ── Roadmap ──────────────────────────────────────────────────────────
\chapter*{Roadmap}\label{ch:roadmap}
\addcontentsline{toc}{chapter}{Roadmap}

This guide develops the mathematical background for the thesis
\emph{Spectral Methods for Binary Neural Networks} in seven chapters,
each building on the previous ones.

\begin{description}[style=nextline]
\item[Chapter~\ref{ch:prereqs}: Prerequisites.]
  Inner product spaces, the Kronecker product, and the probability tools
  (Markov/Chebyshev inequalities, CLT, Berry--Ess\'een) used later.
  Skip this if you are comfortable with undergraduate linear algebra and
  probability.

\item[Chapter~\ref{ch:boolean-fourier}: Fourier Analysis on the Boolean Hypercube.]
  The core of the guide. We define the parity basis, prove orthonormality,
  derive the Fourier expansion and Parseval's identity, and work through
  detailed examples (dictator, AND, majority, parity). We also prove the
  sign-rounding error bound used in the thesis.

\item[Chapter~\ref{ch:influence-noise}: Influence and Noise Sensitivity.]
  Influence of variables, total influence, the noise operator $T_\rho$,
  noise stability, and their spectral formulas. These quantify the
  sensitivity of Boolean functions to perturbations.

\item[Chapter~\ref{ch:concentration}: Hypercontractivity and Spectral Concentration.]
  The Bonami--Beckner inequality, the level-$k$ inequality, and the
  LMN theorem. These are the main tools for proving that ``simple''
  Boolean functions concentrate their spectral energy on low degrees.

\item[Chapter~\ref{ch:wht}: The Walsh--Hadamard Transform.]
  Hadamard matrices, the WHT as the computational realization of the
  Fourier expansion, the fast algorithm (FWHT), and its comparison with
  the FFT.

\item[Chapter~\ref{ch:ltf-codes-groups}: Linear Threshold Functions, Codes, and Groups.]
  Three topics connecting the theory to the thesis: LTFs and Chow's theorem,
  Reed--Muller codes and spectral truncation, and the group-theoretic
  perspective on the Walsh--Fourier basis.

\item[Chapter~\ref{ch:applications}: Application to Binary Neural Networks.]
  How all the preceding material applies to the thesis: the spectral
  parameterization, normalization, and a chapter-by-chapter map from
  this guide to the thesis.
\end{description}

\noindent
\textbf{Dependencies.} Chapters~1--2 are foundational and should be
read first. Chapter~3 depends on~2. Chapter~4 depends on~3.
Chapter~5 depends on~1--2 but is largely independent of~3--4.
Chapter~6 depends on~2 and~5. Chapter~7 draws on all preceding chapters.
See also the dependency graph in \Cref{sec:dependency}.

\newpage

\input{ch1_prerequisites}
\input{ch2_boolean_fourier}
\input{ch3_influence_noise}
\input{ch4_concentration}
\input{ch5_wht}
\input{ch6_ltf_codes_groups}
\input{ch7_applications}

\bibliographystyle{plainnat}
\bibliography{guide_references}

% ── Index ────────────────────────────────────────────────────────────
\cleardoublepage
\addcontentsline{toc}{chapter}{Index}
\printindex


% ── Back-of-book blurb ──────────────────────────────────────────────
\cleardoublepage
\thispagestyle{empty}
\vspace*{\fill}

\begin{center}
{\Large\bfseries About This Guide}
\end{center}

\vspace{0.8cm}

\noindent
\textit{Boolean Fourier Analysis and the Walsh--Hadamard Transform}
provides a self-contained, proof-complete exposition of the mathematics
behind spectral methods for binary neural networks. Starting from
undergraduate linear algebra and probability, it develops the Fourier
analysis of Boolean functions on the hypercube $\{-1,+1\}^n$, covers
influence theory, noise sensitivity, and hypercontractivity, presents the
fast Walsh--Hadamard transform, and explores connections to Reed--Muller
codes, linear threshold functions, and the representation theory of
$\mathbb{Z}_2^n$. The final chapter traces every concept back to its
application in training neural networks with binary weights.

\vspace{0.5cm}

\noindent
\textbf{Features:}
\begin{itemize}[nosep,leftmargin=1.2em]
  \item Full proofs and worked examples for every major result.
  \item A companion Jupyter notebook with NumPy and PyTorch
    implementations of all key algorithms.
  \item Chapter-by-chapter map linking the mathematical theory to the
    thesis on spectral methods for binary neural networks.
  \item Concept dependency graph and curated reading lists.
\end{itemize}

\vspace{0.5cm}

\noindent
\textit{Written as a companion to the thesis} Spectral Methods for Binary
Neural Networks, \textit{this guide serves both as a reference for
researchers at the intersection of Boolean analysis and machine learning,
and as a self-study introduction for students encountering these ideas
for the first time.}

\vspace*{\fill}

\begin{center}
\begin{tikzpicture}[scale=0.45]
  \pgfmathsetmacro{\phi}{0.618}
  \pgfmathsetmacro{\sc}{1.8}
  \foreach \i in {0,...,15} {
    \pgfmathsetmacro{\va}{1-2*Mod(\i,2)}
    \pgfmathsetmacro{\vb}{1-2*Mod(int(\i/2),2)}
    \pgfmathsetmacro{\vc}{1-2*Mod(int(\i/4),2)}
    \pgfmathsetmacro{\vd}{1-2*Mod(int(\i/8),2)}
    \pgfmathsetmacro{\px}{\sc*(\va + \phi*\vc)}
    \pgfmathsetmacro{\py}{\sc*(\vb + \phi*\vd)}
    \coordinate (V\i) at (\px, \py);
  }
  \foreach \i in {0,2,4,6,8,10,12,14} {
    \pgfmathtruncatemacro{\j}{\i+1}
    \draw[black!30, line width=0.8pt] (V\i) -- (V\j);
  }
  \foreach \i in {0,1,4,5,8,9,12,13} {
    \pgfmathtruncatemacro{\j}{\i+2}
    \draw[black!30, line width=0.8pt] (V\i) -- (V\j);
  }
  \foreach \i in {0,1,2,3,8,9,10,11} {
    \pgfmathtruncatemacro{\j}{\i+4}
    \draw[black!30, line width=0.8pt] (V\i) -- (V\j);
  }
  \foreach \i in {0,...,7} {
    \pgfmathtruncatemacro{\j}{\i+8}
    \draw[black!30, line width=0.8pt] (V\i) -- (V\j);
  }
  \foreach \i in {0,...,15} {
    \fill[black!40] (V\i) circle (1.8pt);
  }
\end{tikzpicture}
\end{center}

\end{document}
