\begin{thebibliography}{99}

\bibitem{ArnoldEtAl2024}
N.~Arnold, L.~Holtorf, F.~Sch\"afer, and F.~L\"orch.
\newblock Phase transitions in large language models and the Ising model.
\newblock \emph{arXiv preprint}, arXiv:2405.17088, 2024.

\bibitem{BalleEtAl2024}
B.~Balle, C.~Lacroce, P.~Panangaden, D.~Precup, and G.~Rabusseau.
\newblock Extracting weighted finite automata from recurrent neural networks: multi-letter case.
\newblock Manuscript, 2024.

\bibitem{BanderierDrmota2015}
C.~Banderier and M.~Drmota.
\newblock Formulae and asymptotics for coefficients of algebraic functions.
\newblock \emph{Combinatorics, Probability and Computing}, 24(1):1--53, 2015.

\bibitem{BanderierKubaWallner2021}
C.~Banderier, M.~Kuba, S.~Wagner, and M.~Wallner.
\newblock Phase transitions of composition schemes: {M}ittag-{L}effler and mixed {P}oisson distributions.
\newblock \emph{Ann.\ Appl.\ Probab.}, 34(5):4500--4546, 2024.
\newblock (arXiv:2103.03751, 2021.)

\bibitem{Berman2025}
V.~Berman.
\newblock Random text, {Z}ipf's law, critical length, and implications for large language models.
\newblock \emph{arXiv preprint}, arXiv:2511.17575, 2025.

\bibitem{BerstelReutenauer2011}
J.~Berstel and C.~Reutenauer.
\newblock \emph{Noncommutative Rational Series with Applications}.
\newblock Cambridge University Press, 2011.

\bibitem{BlondelEtAl2025}
M.~Blondel, M.~E. Sander, G.~Vivier-Ardisson, T.~Liu, and V.~Roulet.
\newblock Autoregressive language models are secretly energy-based models: insights into the lookahead capabilities of next-token prediction.
\newblock \emph{arXiv preprint}, arXiv:2512.15605, 2025.

\bibitem{BoothThompson1973}
T.~L. Booth and R.~A. Thompson.
\newblock Applying probability measures to abstract languages.
\newblock \emph{IEEE Trans.\ Comput.}, C-22(5):442--450, 1973.

\bibitem{Bowen1975}
R.~Bowen.
\newblock \emph{Equilibrium States and the Ergodic Theory of Anosov Diffeomorphisms}.
\newblock Lecture Notes in Math.\ vol.~470, Springer, 1975.

\bibitem{Chi1999}
Z.~Chi.
\newblock Statistical properties of probabilistic context-free grammars.
\newblock \emph{Computational Linguistics}, 25(1):131--160, 1999.

\bibitem{ChomskySchutzenberger1963}
N.~Chomsky and M.~P. Sch\"utzenberger.
\newblock The algebraic theory of context-free languages.
\newblock In \emph{Computer Programming and Formal Systems}, pages 118--161. North-Holland, 1963.

\bibitem{CotterellEtAl2023}
R.~Cotterell, A.~Svete, C.~Meister, and others.
\newblock Formal language theory meets modern NLP.
\newblock Survey/tutorial, 2023.

\bibitem{Drmota2009}
M.~Drmota.
\newblock \emph{Random Trees}.
\newblock Springer, Vienna, 2009.

\bibitem{DrosteKuichVogler2009}
M.~Droste, W.~Kuich, and H.~Vogler, editors.
\newblock \emph{Handbook of Weighted Automata}.
\newblock Springer, 2009.

\bibitem{DuEtAl2023}
L.~Du, L.~Hennigen, T.~Pimentel, C.~Meister, J.~Eisner, and R.~Cotterell.
\newblock A measure-theoretic characterization of tight language models.
\newblock In \emph{Proc.\ ACL}, 2023.

\bibitem{DuchonFlajoletLouchardSchaeffer2004}
P.~Duchon, P.~Flajolet, G.~Louchard, and G.~Schaeffer.
\newblock Boltzmann samplers for the random generation of combinatorial structures.
\newblock \emph{Combin.\ Probab.\ Comput.}, 13(4--5):577--625, 2004.

\bibitem{FlajoletOdlyzko1990}
P.~Flajolet and A.~M. Odlyzko.
\newblock Singularity analysis of generating functions.
\newblock \emph{SIAM J.\ Discrete Math.}, 3(2):216--240, 1990.

\bibitem{FlajoletSedgewick2009}
P.~Flajolet and R.~Sedgewick.
\newblock \emph{Analytic Combinatorics}.
\newblock Cambridge University Press, 2009.

\bibitem{FlajoletZimmermannVanCutsem1994}
P.~Flajolet, P.~Zimmermann, and B.~Van~Cutsem.
\newblock A calculus for the random generation of labelled combinatorial structures.
\newblock \emph{Theoret.\ Comput.\ Sci.}, 132(1--2):1--35, 1994.

\bibitem{GoldwaterGriffithsJohnson2011}
S.~Goldwater, T.~L. Griffiths, and M.~Johnson.
\newblock Producing power-law distributions and damping word frequencies with two-stage language models.
\newblock \emph{J.\ Mach.\ Learn.\ Res.}, 12:2335--2382, 2011.

\bibitem{GuibasOdlyzko1981}
L.~J. Guibas and A.~M. Odlyzko.
\newblock String overlaps, pattern matching, and nontransitive games.
\newblock \emph{J.\ Combin.\ Theory Ser.~A}, 30(2):183--208, 1981.

\bibitem{Hahn2020}
M.~Hahn.
\newblock Theoretical limitations of self-attention in neural sequence models.
\newblock \emph{Trans.\ Assoc.\ Comput.\ Linguist.}, 8:156--171, 2020.

\bibitem{HopcroftUllman1979}
J.~E. Hopcroft and J.~D. Ullman.
\newblock \emph{Introduction to Automata Theory, Languages, and Computation}.
\newblock Addison-Wesley, 1979.

\bibitem{Hwang1998}
H.-K. Hwang.
\newblock On convergence rates in the central limit theorems for combinatorial structures.
\newblock \emph{European J.\ Combin.}, 19(3):329--343, 1998.

\bibitem{Icard2020}
T.~Icard.
\newblock Calibrating generative models: the probabilistic {C}homsky--{S}ch\"utzenberger hierarchy.
\newblock \emph{J.\ Math.\ Psych.}, 95:102308, 2020.

\bibitem{JacquetSzpankowski1991}
P.~Jacquet and W.~Szpankowski.
\newblock Analysis of digital tries with {M}arkovian dependency.
\newblock \emph{IEEE Trans.\ Inform.\ Theory}, 37(5):1470--1475, 1991.

\bibitem{KemptonBurrell2025}
T.~Kempton and C.~Burrell.
\newblock Temperature and local normalization in autoregressive models.
\newblock Manuscript, 2025.

\bibitem{KontoyiannisEtAl1998}
I.~Kontoyiannis, P.~H. Algoet, Y.~M. Suhov, and A.~J. Barron.
\newblock Nonparametric entropy estimation for stationary processes and random fields.
\newblock \emph{IEEE Trans.\ Inform.\ Theory}, 44(3):1319--1327, 1998.

\bibitem{LacroceEtAl2024}
C.~Lacroce, B.~Balle, P.~Panangaden, and G.~Rabusseau.
\newblock Extracting weighted finite automata from recurrent neural networks via the one-letter setting.
\newblock \emph{Math.\ Structures Comput.\ Sci.}, 2024.

\bibitem{LacrocePanangadenRabusseau2021}
C.~Lacroce, P.~Panangaden, and G.~Rabusseau.
\newblock Extracting weighted finite automata from recurrent neural networks using constrained {AAK} theory.
\newblock In \emph{Proc.\ ICGI}, 2021.

\bibitem{LecorveMotlicek2012}
G.~Lecorv\'e and P.~Motlicek.
\newblock Conversion of recurrent neural network language models to weighted finite-state transducers for automatic speech recognition.
\newblock In \emph{Proc.\ Interspeech}, 2012.

\bibitem{LiEtAl2024}
Z.~Li and collaborators.
\newblock Finite-size scaling analysis of language models.
\newblock \emph{arXiv preprint}, 2024.

\bibitem{LinTegmarkRolnick2017}
H.~W. Lin, M.~Tegmark, and D.~Rolnick.
\newblock Why does deep and cheap learning work so well?
\newblock \emph{J.\ Statist.\ Phys.}, 168:1223--1247, 2017.

\bibitem{MerrillSabharwal2024}
W.~Merrill and A.~Sabharwal.
\newblock The expressive power of transformers with chain of thought.
\newblock In \emph{Proc.\ ICLR}, 2024.

\bibitem{MerrillSabharwalSmith2022}
W.~Merrill, A.~Sabharwal, and N.~A. Smith.
\newblock Saturated transformers are constant-depth threshold circuits.
\newblock \emph{Trans.\ Assoc.\ Comput.\ Linguist.}, 10:843--856, 2022.

\bibitem{MichaudEtAl2023}
E.~J. Michaud, Z.~Liu, U.~Girit, and M.~Tegmark.
\newblock The quantization model of neural scaling laws.
\newblock In \emph{Proc.\ NeurIPS}, arXiv:2303.13506, 2023.

\bibitem{Mikhaylovskiy2025}
N.~Mikhaylovskiy.
\newblock Universality of phase transitions in large language models.
\newblock \emph{Findings of EMNLP}, 2025.

\bibitem{Mikhaylovskiy2025b}
N.~Mikhaylovskiy.
\newblock {Z}ipf's law and the temperature window in transformer language models.
\newblock \emph{Findings of EMNLP}, findings-emnlp.837, 2025.

\bibitem{NakaishiNishikawaHukushima2024}
Y.~Nakaishi, Y.~Nishikawa, and K.~Hukushima.
\newblock Phase transitions in {GPT-2}: a statistical mechanical analysis.
\newblock \emph{arXiv preprint}, arXiv:2406.05335, 2024.

\bibitem{NicodemeSalvyFlajolet2002}
P.~Nicod\`eme, B.~Salvy, and P.~Flajolet.
\newblock Motif statistics.
\newblock \emph{Theoret.\ Comput.\ Sci.}, 287(2):593--617, 2002.

\bibitem{NijenhuisWilf1978}
A.~Nijenhuis and H.~S. Wilf.
\newblock \emph{Combinatorial Algorithms}.
\newblock Academic Press, 2nd edition, 1978.

\bibitem{PemantleWilson2013}
R.~Pemantle and M.~C. Wilson.
\newblock \emph{Analytic Combinatorics in Several Variables}.
\newblock Cambridge University Press, 2013.

\bibitem{PengEtAl2018}
H.~Peng, R.~Schwartz, S.~Thomson, and N.~A. Smith.
\newblock Rational recurrences.
\newblock In \emph{Proc.\ EMNLP}, 2018.

\bibitem{Piantadosi2014}
S.~T. Piantadosi.
\newblock {Z}ipf's word frequency law in natural language: a critical review and future directions.
\newblock \emph{Psychon.\ Bull.\ Rev.}, 21(5):1112--1130, 2014.

\bibitem{RabusseauLiPrecup2019}
G.~Rabusseau, T.~Li, and D.~Precup.
\newblock Connecting weighted automata and recurrent neural networks through spectral learning.
\newblock In \emph{Proc.\ AISTATS}, 2019.

\bibitem{RizviEtAl2024}
M.~Rizvi-Martel, M.~Lizaire, C.~Lacroce, and G.~Rabusseau.
\newblock Simulating weighted automata over sequences and trees with transformers.
\newblock In \emph{Proc.\ AISTATS}, pages 2368--2376, 2024.

\bibitem{Ruelle1978}
D.~Ruelle.
\newblock \emph{Thermodynamic Formalism}.
\newblock Addison-Wesley, 1978.

\bibitem{ScheibnerSmithBialek2025}
D.~Scheibner, C.~Smith, and W.~Bialek.
\newblock Entropy rate of natural language revisited.
\newblock \emph{arXiv preprint}, arXiv:2512.24969, 2025.

\bibitem{SchwartzThomsonSmith2018}
R.~Schwartz, S.~Thomson, and N.~A. Smith.
\newblock Bridging {CNN}s, {RNN}s, and weighted finite-state machines.
\newblock In \emph{Proc.\ ACL}, 2018.

\bibitem{Shannon1951}
C.~E. Shannon.
\newblock Prediction and entropy of printed {E}nglish.
\newblock \emph{Bell Syst.\ Tech.\ J.}, 30(1):50--64, 1951.

\bibitem{Sipser2013}
M.~Sipser.
\newblock \emph{Introduction to the Theory of Computation}.
\newblock Cengage, 3rd edition, 2013.

\bibitem{SmithJohnson2007}
N.~A. Smith and M.~Johnson.
\newblock Weighted and probabilistic context-free grammars are equally expressive.
\newblock \emph{Comput.\ Linguist.}, 33(4):477--491, 2007.

\bibitem{StroblEtAl2024}
L.~Strobl, W.~Merrill, G.~Weiss, D.~Chiang, and D.~Angluin.
\newblock What formal languages can transformers express? {A} survey.
\newblock \emph{Trans.\ Assoc.\ Comput.\ Linguist.}, 12:543--561, 2024.

\bibitem{SureshEtAl2021}
K.~Suresh, B.~Roark, M.~Riley, and V.~Schogol.
\newblock Distilling a neural network language model into a finite-state transducer for on-device ASR.
\newblock In \emph{Proc.\ Interspeech}, 2021.

\bibitem{SveteCotterell2023}
A.~Svete and R.~Cotterell.
\newblock Recurrent neural networks and weighted finite-state transducers.
\newblock In \emph{Proc.\ ACL}, 2023.

\bibitem{Szpankowski2001}
W.~Szpankowski.
\newblock \emph{Average Case Analysis of Algorithms on Sequences}.
\newblock Wiley, 2001.

\bibitem{TakahashiTanakaIshii2018}
S.~Takahashi and K.~Tanaka-Ishii.
\newblock Cross entropy of neural language models at infinity.
\newblock In \emph{Proc.\ EMNLP}, 2018.

\bibitem{VaswaniEtAl2017}
A.~Vaswani, N.~Shazeer, N.~Parmar, J.~Uszkoreit, L.~Jones, A.~N. Gomez, {\L}.~Kaiser, and I.~Polosukhin.
\newblock Attention is all you need.
\newblock In \emph{Proc.\ NeurIPS}, 2017.

\bibitem{WeissGoldbergYahav2018}
G.~Weiss, Y.~Goldberg, and E.~Yahav.
\newblock Extracting automata from recurrent neural networks using queries and counterexamples.
\newblock In \emph{Proc.\ ICML}, 2018.

\bibitem{YunEtAl2020}
C.~Yun, S.~Bhojanapalli, A.~S. Rawat, S.~Reddi, and S.~Kumar.
\newblock Are transformers universal approximators of sequence-to-sequence functions?
\newblock In \emph{Proc.\ ICLR}, 2020.

\bibitem{Zipf1949}
G.~K. Zipf.
\newblock \emph{Human Behavior and the Principle of Least Effort}.
\newblock Addison-Wesley, 1949.

\end{thebibliography}
