\documentclass[11pt,a4paper]{book}
\input{preamble}

\title{\Huge Analytic Combinatorics and\\ Large Language Models\\[0.5em]
       \Large An Expository Introduction}
\author{RACA Notes}
\date{April 2026}

\begin{document}

\frontmatter
\maketitle

\tableofcontents

\chapter*{Notation}
\addcontentsline{toc}{chapter}{Notation}

\begin{tabular}{@{}ll@{}}
\textbf{Symbol} & \textbf{Meaning} \\[4pt]
\multicolumn{2}{@{}l}{\emph{Generating functions and asymptotics}} \\
$A(z), B(z), C(z), \ldots$ & Ordinary generating functions (OGFs) \\
$\hat{A}(z)$ & Exponential generating function (EGF) of $A$ \\
$[z^n] A(z)$ & Coefficient of $z^n$ in $A(z)$ \\
$[z^n/n!] \hat{A}(z)$ & $n! \cdot [z^n] \hat{A}(z)$; the $n$-th EGF coefficient \\
$R$ & Radius of convergence (Cauchy--Hadamard) \\
$\rho$ & Location of dominant singularity; equals $R$ for aperiodic GFs \\
$\alpha$ & Singularity exponent in $(1-z/\rho)^{-\alpha}$ \\
$f(n) \sim g(n)$ & $\lim_{n\to\infty} f(n)/g(n) = 1$ \\[6pt]

\multicolumn{2}{@{}l}{\emph{Combinatorial constructions}} \\
$\mathcal{E}$ & Neutral class (one object of size $0$; OGF $= 1$) \\
$\mathcal{Z}$ & Atomic class (one object of size $1$; OGF $= z$) \\
$\SEQ(\mathcal{A})$ & Sequences of $\mathcal{A}$-objects; OGF $= 1/(1-A(z))$ \\
$\SET(\mathcal{A})$ & Unordered sets; EGF $= \exp\hat{A}(z)$ \\
$\CYC(\mathcal{A})$ & Cyclic sequences; EGF $= \log(1/(1-\hat{A}(z)))$ \\[6pt]

\multicolumn{2}{@{}l}{\emph{Formal languages and automata}} \\
$\Sigma$ & Finite alphabet \\
$\Sigma^*$ & Set of all finite strings over $\Sigma$ \\
$\EOS$ & End-of-sequence token \\
$L \subseteq \Sigma^*$ & A formal language \\
$(\alpha, \{A_\sigma\}, \beta)$ & Weighted finite automaton: initial vector, transition matrices, final vector \\
$A = \sum_\sigma A_\sigma$ & Aggregate transition matrix of a WFA \\
$\rho(A)$ & Spectral radius of matrix $A$ \\[6pt]

\multicolumn{2}{@{}l}{\emph{Probability and information theory}} \\
$\mu$ & Probability distribution on $\Sigma^*$ (LLM output distribution) \\
$p(\sigma \mid h)$ & Conditional next-token probability given history $h$ \\
$h$ & Shannon entropy rate (bits or nats per symbol) \\
$H(X_1^n)$ & Block entropy of a length-$n$ string \\[6pt]

\multicolumn{2}{@{}l}{\emph{Statistical mechanics / temperature}} \\
$E(w) = -\log \mu(w)$ & Energy of string $w$ under model $\mu$ \\
$\beta = 1/T$ & Inverse temperature \\
$Z(\beta) = \sum_w \mu(w)^\beta$ & Partition function \\
$\mu_\beta(w) \propto \mu(w)^\beta$ & Gibbs distribution at inverse temperature $\beta$ \\
$\beta_c$ & Critical inverse temperature (phase transition point) \\[6pt]

\multicolumn{2}{@{}l}{\emph{Named constants}} \\
$\varphi = (1+\sqrt{5})/2$ & Golden ratio ($\approx 1.618$) \\
$\hat\varphi = (1-\sqrt{5})/2$ & Conjugate of the golden ratio ($\approx -0.618$) \\
$\Gamma(\alpha)$ & Euler gamma function \\
\end{tabular}

\bigskip

\noindent Throughout, $\log$ denotes the natural logarithm unless a base is
indicated (as in $\log_2$).

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

This book is an extended expository introduction to the research territory
where analytic combinatorics meets large language models. It is written for a
student who knows calculus, linear algebra, and a little probability, has
perhaps seen complex numbers, and wants to understand both halves of the
subject from scratch: the symbolic method and singularity analysis of Flajolet
and Sedgewick on one side, and the probabilistic, formal-language, and
statistical-mechanics views of neural language models on the other.

The organizing claim of the book is that these two traditions, despite rarely
citing each other, are talking about the same mathematical objects: partition
functions are generating functions, critical temperatures are radii of
convergence, phase transitions are singularities, rational-series theorems
from automata are dominant-pole theorems from analytic combinatorics. The
research frontier consists largely of building bridges that the two communities
have, until recently, avoided building. The final chapters make those bridges
explicit and give a student a plausible starting point.

Part~I develops the classical machinery of analytic combinatorics from zero,
up to the $n^{-3/2}\rho^{-n}$ asymptotic for unambiguous context-free
languages. Part~II turns to weighted formal languages, probabilistic grammars,
tightness, and the entropy-rate identity $h = \log(1/R)$. Part~III introduces
autoregressive language models and the formal-language theory of transformers,
then surveys the weighted-finite-automaton approximation pipeline. Part~IV is
devoted to Boltzmann sampling, Gibbs-measure views of temperature decoding, and
phase transitions. Part~V covers Zipf's law, motif statistics, and the ten open
problems we currently see at the frontier.

Exercises appear in selected chapters but are not the point. The point is to get
a reader who starts knowing none of this to the point where they can open an
arXiv paper from 2025 on, say, phase transitions in Pythia or WFA extraction
from a Mistral model, and see what is going on.

\mainmatter
\input{partI}
\input{partII}
\input{partIII}
\input{partIV}
\input{partV}

\backmatter
\input{bibliography}

\end{document}
