\chapter{What is a Generating Function?}

A sequence of numbers $(a_n)_{n \ge 0}$ is the natural output of a counting problem: how many binary strings of length $n$, how many trees with $n$ nodes, how many ways to parse a sentence of length $n$ according to a given grammar. Studying the sequence directly forces us to deal with each index separately, and recurrences --- while precise --- often resist closed-form extraction. The central idea of generating function theory is to encode the entire sequence as a single algebraic object,
\[
  A(z) \;=\; \sum_{n \ge 0} a_n z^n,
\]
and then let the algebra of formal power series do the bookkeeping. This is not mere notational convenience. Operations on sequences that would otherwise require index gymnastics correspond to clean algebraic operations on $A(z)$:
\[
  a_{n-1} \longleftrightarrow zA(z), \qquad
  \sum_{k=0}^n a_k b_{n-k} \longleftrightarrow A(z)B(z), \qquad
  a_n-a_{n-1} \longleftrightarrow (1-z)A(z)
\]
after the appropriate initial-term corrections. Recurrences, which are linear or polynomial relations among the $a_n$, become \emph{functional equations} for $A(z)$, and those equations can often be solved in closed form. The solution then exposes structure invisible in the raw sequence.

\begin{definition}
The set of \emph{formal power series} over $\mathbb{Q}$, denoted $\mathbb{Q}[[z]]$, consists of all expressions $A(z) = \sum_{n \ge 0} a_n z^n$ with $a_n \in \mathbb{Q}$. Two series are added term by term: $(A+B)(z) = \sum_n (a_n + b_n) z^n$. Their product is defined by
\[
  (A \cdot B)(z) \;=\; \sum_{n \ge 0} c_n z^n, \qquad c_n = \sum_{k=0}^{n} a_k b_{n-k}.
\]
This product is well-defined because the coefficient of $z^n$ depends on only finitely many pairs $(a_k,b_{n-k})$, namely $k=0,1,\dots,n$.
\end{definition}

The product formula follows from collecting all pairs of terms $a_k z^k$ and $b_m z^m$ whose exponents add to $n$: $z^k \cdot z^m = z^n$ requires $m = n-k$, so the coefficient of $z^n$ in the product is $a_0 b_n + a_1 b_{n-1} + \cdots + a_n b_0$. At this stage $z$ is a \emph{formal indeterminate} --- it carries no numerical value, and questions of convergence are entirely irrelevant. In $\mathbb{Q}[[z]]$ one can always add, subtract, and multiply series, and the proposition below will show exactly when division is possible. (Algebraists call $\mathbb{Q}[[z]]$ a \emph{commutative ring}; this terminology is not needed for what follows.)

\begin{definition}
The \emph{coefficient extraction operator} $[z^n]$ picks out the $n$-th coefficient of a power series:
\[
  [z^n] A(z) \;:=\; a_n.
\]
It is linear because addition in $\mathbb{Q}[[z]]$ is term by term:
\[
  [z^n](A+B) = [z^n]\sum_{m \ge 0}(a_m+b_m)z^m = a_n+b_n = [z^n]A + [z^n]B.
\]
For products, one simply reads off the coefficient of $z^n$ in the Cauchy product:
\[
  [z^n](A \cdot B) = \sum_{k=0}^{n} a_k b_{n-k}.
\]
\end{definition}

\begin{proposition}
Let $A(z) = \sum_{n \ge 0} a_n z^n \in \mathbb{Q}[[z]]$. Then $A(z)$ has a multiplicative inverse in $\mathbb{Q}[[z]]$ if and only if $a_0 \ne 0$.
\end{proposition}

\begin{proof}
Suppose first that $a_0 \ne 0$. We seek a series $D(z) = \sum_{n \ge 0} d_n z^n$ such that
\[
  A(z)D(z) = 1.
\]
Comparing coefficients of equal powers of $z$ determines the $d_n$ recursively. The constant term gives
\[
  a_0 d_0 = 1,
\]
so $d_0 = 1/a_0$. For $n \ge 1$, the coefficient of $z^n$ in $A(z)D(z)$ must be $0$, so
\[
  a_0 d_n + a_1 d_{n-1} + \cdots + a_n d_0 = 0.
\]
Therefore
\[
  d_n = -\frac{a_1 d_{n-1} + a_2 d_{n-2} + \cdots + a_n d_0}{a_0}.
\]
Each $d_n$ is uniquely determined by the earlier coefficients $d_0,\dots,d_{n-1}$, so this recursion produces one and only one series $D(z)$ with $A(z)D(z)=1$.

Conversely, if $a_0 = 0$, then the constant term of $A(z)D(z)$ is always $a_0d_0 = 0$ for every series $D(z)$. Such a product can never equal the constant series $1$. So no inverse exists in that case.
\end{proof}

\begin{lemma}
In the ring of formal power series,
\[
  \frac{1}{1-x} \;=\; \sum_{n \ge 0} x^n.
\]
\end{lemma}

\begin{proof}
Let $G(x) = \sum_{n \ge 0} x^n$. Then
\[
  (1-x)G(x)
  = \sum_{n \ge 0} x^n - \sum_{n \ge 0} x^{n+1}
  = 1 + \sum_{n \ge 1}(1-1)x^n
  = 1.
\]
So $G(x)$ is the multiplicative inverse of $1-x$.
\end{proof}

Two operations on power series appear repeatedly. The \emph{shift} by $z$ sends
\[
  A(z) \longmapsto zA(z) = \sum_{n \ge 1} a_{n-1} z^n,
\]
so $[z^n](zA(z)) = a_{n-1}$. Likewise,
\[
  (1-z)A(z) = a_0 + \sum_{n \ge 1}(a_n-a_{n-1})z^n,
\]
so multiplying by $1-z$ encodes first differences after the constant term. The \emph{formal derivative} is
\[
  A'(z) = \sum_{n \ge 1} n a_n z^{n-1},
\]
defined term by term with no analytic content. It still satisfies the usual product rule: if $f(z)=\sum a_n z^n$ and $g(z)=\sum b_n z^n$, then
\[
  [z^{n-1}](fg)' = n\sum_{k=0}^n a_k b_{n-k},
\]
while
\[
  [z^{n-1}](f'g+fg')
  = \sum_{k=1}^n k a_k b_{n-k} + \sum_{k=0}^{n-1} (n-k)a_k b_{n-k}
  = n\sum_{k=0}^n a_k b_{n-k},
\]
so $(fg)' = f'g + fg'$ coefficient by coefficient.

\section*{The method: from recurrences to functional equations}

The central technique of this chapter is a three-step recipe for converting a recurrence into a closed-form generating function.

\begin{enumerate}
\item \textbf{Define} the generating function $A(z) = \sum_{n \ge 0} a_n z^n$.
\item \textbf{Multiply} every term of the recurrence $a_n = (\text{expression in } a_{n-1}, a_{n-2}, \ldots)$ by $z^n$ and \textbf{sum} over all valid $n$. Recognize each resulting sum as an expression involving $A(z)$.
\item \textbf{Solve} the resulting equation for $A(z)$.
\end{enumerate}

Step~2 is the key move. The slogan is that $a_{n-k}$ turns into $z^kA(z)$, but one must watch the lower limit of summation. The clean identity is
\[
  \sum_{n \ge k} a_{n-k} z^n = z^k A(z).
\]
More generally, if the sum begins at $n=N$, then after the substitution $m=n-k$ one gets
\[
  \sum_{n \ge N} a_{n-k} z^n
  = z^k \sum_{m \ge N-k} a_m z^m
  = z^k\bigl(A(z) - a_0 - a_1 z - \cdots - a_{N-k-1}z^{N-k-1}\bigr).
\]
So the only nuisance is bookkeeping for the missing initial terms. Once those are handled, the recurrence --- an infinite family of equations, one for each $n$ --- collapses into a single equation for the single unknown $A(z)$.

We now demonstrate this on three progressively harder examples.

\section*{Binary strings}

Let $b_n$ denote the number of binary strings of length $n$. A direct count already gives $b_n = 2^n$, since each of the $n$ positions can be filled by either $0$ or $1$. We now re-derive the same answer by generating functions in order to illustrate the method. Any binary string of length $n \ge 1$ is obtained by appending one of two symbols to a string of length $n-1$, giving
\[
  b_n = 2b_{n-1} \qquad (n \ge 1), \qquad b_0 = 1.
\]

Define $B(z) = \sum_{n \ge 0} b_n z^n$, multiply the recurrence by $z^n$, and sum over $n \ge 1$:
\[
  \sum_{n \ge 1} b_n z^n \;=\; 2z \sum_{n \ge 1} b_{n-1} z^{n-1}.
\]
The left side is $B(z)-b_0 = B(z)-1$. For the right side, let $m=n-1$. Then $m \ge 0$, so
\[
  \sum_{n \ge 1} b_{n-1} z^{n-1}
  = \sum_{m \ge 0} b_m z^m
  = B(z).
\]
Thus
\[
  B(z) - 1 = 2zB(z), \qquad \text{so} \qquad B(z) = \frac{1}{1-2z}.
\]
By the formal geometric-series lemma,
\[
  \frac{1}{1-2z} = \sum_{n \ge 0} (2z)^n = \sum_{n \ge 0} 2^n z^n,
\]
so $[z^n]B(z)=2^n$ as expected. The functional-equation route is more instructive than the direct count because it still works when no closed form is obvious beforehand.

\section*{Fibonacci numbers}

The Fibonacci sequence is defined by $F_0 = 0$, $F_1 = 1$, and
\[
  F_n = F_{n-1} + F_{n-2} \qquad (n \ge 2).
\]
Set $F(z) = \sum_{n \ge 0} F_n z^n$. Multiply the recurrence by $z^n$ and sum over $n \ge 2$:
\[
  \sum_{n \ge 2} F_n z^n
  \;=\; z \sum_{n \ge 2} F_{n-1} z^{n-1}
  \;+\; z^2 \sum_{n \ge 2} F_{n-2} z^{n-2}.
\]
The left side is
\[
  \sum_{n \ge 2} F_n z^n = F(z) - F_0 - F_1 z = F(z)-z.
\]
For the first sum on the right, let $m=n-1$. Then $m \ge 1$, so
\[
  \sum_{n \ge 2} F_{n-1} z^{n-1}
  = \sum_{m \ge 1} F_m z^m
  = F(z)-F_0
  = F(z).
\]
For the second, let $m=n-2$. Then $m \ge 0$, so
\[
  \sum_{n \ge 2} F_{n-2} z^{n-2}
  = \sum_{m \ge 0} F_m z^m
  = F(z).
\]
Therefore
\[
  F(z)-z = zF(z) + z^2F(z),
\]
and hence
\[
  F(z) = \frac{z}{1-z-z^2}.
\]

To extract $F_n$ explicitly, we factor the denominator and use partial fractions. The roots of $x^2-x-1=0$ are
\[
  x = \frac{1 \pm \sqrt{5}}{2},
\]
so we define
\[
  \varphi = \frac{1+\sqrt{5}}{2}, \qquad \hat{\varphi} = \frac{1-\sqrt{5}}{2}.
\]
Then $\varphi+\hat{\varphi}=1$ and $\varphi\hat{\varphi}=-1$, so
\[
  1-z-z^2 = (1-\varphi z)(1-\hat{\varphi} z).
\]
Because the denominator has two distinct linear factors, elementary algebra tells us that a proper rational function with this denominator can be written in the form
\[
  \frac{A}{1-\varphi z} + \frac{B}{1-\hat{\varphi} z}.
\]
So we seek constants $A$ and $B$ such that
\[
  F(z)
  = \frac{z}{(1-\varphi z)(1-\hat{\varphi} z)}
  = \frac{A}{1-\varphi z} + \frac{B}{1-\hat{\varphi} z}.
\]
Multiply both sides by $(1-\varphi z)(1-\hat{\varphi} z)$:
\[
  z = A(1-\hat{\varphi} z) + B(1-\varphi z).
\]
Comparing constant terms gives $A+B=0$, so $B=-A$. Comparing coefficients of $z$ gives
\[
  1 = -A\hat{\varphi} - B\varphi = A(\varphi-\hat{\varphi}) = A\sqrt{5},
\]
so $A=1/\sqrt{5}$ and $B=-1/\sqrt{5}$. Therefore
\[
  F(z)
  = \frac{1}{\sqrt{5}} \cdot \frac{1}{1-\varphi z}
  - \frac{1}{\sqrt{5}} \cdot \frac{1}{1-\hat{\varphi} z}.
\]
Applying the geometric-series lemma to each factor,
\[
  \frac{1}{1-\varphi z} = \sum_{n \ge 0} \varphi^n z^n,
  \qquad
  \frac{1}{1-\hat{\varphi} z} = \sum_{n \ge 0} \hat{\varphi}^n z^n.
\]
Reading off the coefficient of $z^n$ gives Binet's formula:
\[
  F_n = \frac{\varphi^n-\hat{\varphi}^n}{\sqrt{5}}.
\]
Since $|\hat{\varphi}| < 1$, the second term becomes negligible as $n \to \infty$, so
\[
  F_n \sim \frac{\varphi^n}{\sqrt{5}}.
\]
This asymptotic now follows directly from the explicit formula. A deeper preview is that Chapter~4 will recover the same conclusion from the pole of $F(z)$ at $z=1/\varphi$, showing how singularities encode growth rates \cite{FlajoletSedgewick2009}.

\section*{Catalan numbers}

The Catalan numbers $C_n$ count full binary trees (every internal node has exactly two children) with $n$ internal nodes; equivalently, they count triangulations of a convex $(n+2)$-gon, Dyck paths of length $2n$, and well-formed parenthesizations of length $n$, among many other combinatorial families. The sequence begins
\[
  1,\,1,\,2,\,5,\,14,\,42,\dots
\]
and satisfies the recurrence
\[
  C_n = \sum_{k=0}^{n-1} C_k\,C_{n-1-k}, \qquad C_0 = 1.
\]
Here is the structural reason. A full binary tree with $n \ge 1$ internal nodes has a root. If the left subtree has $k$ internal nodes, then the right subtree must have the remaining $n-1-k$. For $n=3$, the three possibilities are $(k,n-1-k)=(0,2),(1,1),(2,0)$.

\begin{center}
\begin{tikzpicture}[
  level distance=0.8cm,
  sibling distance=1.2cm,
  every node/.style={circle, draw, fill=black!10, inner sep=1.5pt},
  leaf/.style={rectangle, draw=none, fill=none, font=\small}
]
  % Generic decomposition
  \node (root) {}
    child {node[draw=none, fill=none] {$\triangle$} edge from parent node[left, draw=none, fill=none] {$k$}}
    child {node[draw=none, fill=none] {$\triangle$} edge from parent node[right, draw=none, fill=none] {$n-1-k$}};
  \node[draw=none, fill=none, right=2cm of root] (eq) {\Large $\longrightarrow$};
  
  % n=3 examples
  \begin{scope}[shift={(5, 0.5)}, sibling distance=0.8cm, level distance=0.6cm]
    \node[draw=none, fill=none, above=0.2cm] {$(0,2)$};
    \node {}
      child {node[leaf] {}}
      child {node {}
        child {node[leaf] {}}
        child {node {}
          child {node[leaf] {}}
          child {node[leaf] {}}
        }
      };
  \end{scope}

  \begin{scope}[shift={(7.5, 0.5)}, sibling distance=0.8cm, level distance=0.6cm]
    \node[draw=none, fill=none, above=0.2cm] {$(1,1)$};
    \node {}
      child {node {}
        child {node[leaf] {}}
        child {node[leaf] {}}
      }
      child {node {}
        child {node[leaf] {}}
        child {node[leaf] {}}
      };
  \end{scope}

  \begin{scope}[shift={(10, 0.5)}, sibling distance=0.8cm, level distance=0.6cm]
    \node[draw=none, fill=none, above=0.2cm] {$(2,0)$};
    \node {}
      child {node {}
        child {node {}
          child {node[leaf] {}}
          child {node[leaf] {}}
        }
        child {node[leaf] {}}
      }
      child {node[leaf] {}};
  \end{scope}
\end{tikzpicture}
\end{center}

Define $C(z)=\sum_{n \ge 0} C_n z^n$. Multiply the recurrence by $z^n$ and sum over $n \ge 1$. The left side is
\[
  \sum_{n \ge 1} C_n z^n = C(z)-1.
\]
For the right side, first expand the square of the series:
\[
  C(z)^2
  = \Bigl(\sum_{i \ge 0} C_i z^i\Bigr)\Bigl(\sum_{j \ge 0} C_j z^j\Bigr)
  = \sum_{m \ge 0}\Bigl(\sum_{k=0}^m C_k C_{m-k}\Bigr) z^m.
\]
So the coefficient of $z^{n-1}$ in $C(z)^2$ is exactly
\[
  [z^{n-1}]\,C(z)^2 = \sum_{k=0}^{n-1} C_k C_{n-1-k}.
\]
Therefore
\[
  \sum_{n \ge 1}\Bigl(\sum_{k=0}^{n-1} C_k C_{n-1-k}\Bigr) z^n
  = z\sum_{n \ge 1} [z^{n-1}]\,C(z)^2\, z^{n-1}
  = z\,C(z)^2.
\]
Equating left and right sides gives
\[
  C(z)-1 = z\,C(z)^2,
\]
or equivalently
\[
  C(z) = 1 + z\,C(z)^2.
\]

This is our first \emph{quadratic} functional equation. We are solving a quadratic equation in the unknown $C(z)$, with coefficients in the field $\mathbb{Q}(z)$. Rearranging,
\[
  z\,C(z)^2 - C(z) + 1 = 0.
\]
Applying the quadratic formula with $a=z$, $b=-1$, $c=1$ yields
\[
  C(z) = \frac{1 \pm \sqrt{1-4z}}{2z}.
\]

\begin{lemma}
There is a unique formal power series
\[
  S(z) = 1 + s_1 z + s_2 z^2 + \cdots
\]
such that $S(z)^2 = 1-4z$. We denote this series by $\sqrt{1-4z}$.
\end{lemma}

\begin{proof}
Write
\[
  S(z)^2 = (1+s_1 z + s_2 z^2 + \cdots)^2.
\]
Comparing constant terms forces $1^2=1$, so the constant term must indeed be $1$. For $n \ge 1$, the coefficient of $z^n$ in $S(z)^2$ has the form
\[
  2s_n + (\text{expression involving only } s_1,\dots,s_{n-1}).
\]
Since the coefficient of $z$ in $1-4z$ is $-4$ and all higher coefficients are $0$, this determines $s_n$ uniquely from the earlier coefficients. So there is exactly one such formal power series.
\end{proof}

Now choose the sign. Since $\sqrt{1-4z}$ has constant term $1$, the numerator
\[
  1 + \sqrt{1-4z}
\]
has constant term $2$, and dividing by $2z$ would produce a $z^{-1}$ term, so it cannot define a formal power series. By contrast,
\[
  1 - \sqrt{1-4z}
\]
has constant term $0$, so division by $2z$ can still produce an honest power series. Therefore
\[
  C(z) = \frac{1-\sqrt{1-4z}}{2z}.
\]

\begin{theorem}[Generalized binomial theorem]
For any rational number $\alpha$,
\[
  (1+w)^\alpha = \sum_{n \ge 0} \binom{\alpha}{n} w^n
\]
as a formal power series, where
\[
  \binom{\alpha}{n}
  = \frac{\alpha(\alpha-1)(\alpha-2)\cdots(\alpha-n+1)}{n!},
  \qquad \binom{\alpha}{0}=1.
\]
This is a formal-power-series statement; no analytic convergence is being used here. We will use the theorem as a black box in this chapter and discuss analytic convergence only in Chapter~2.
\end{theorem}

Applying it with $\alpha=\tfrac12$ and $w=-4z$ gives
\[
  \sqrt{1-4z} = (1-4z)^{1/2} = \sum_{n \ge 0} \binom{1/2}{n}(-4z)^n.
\]
We next simplify the coefficients. For $n \ge 1$,
\[
  \binom{1/2}{n}
  = \frac{\tfrac12(\tfrac12-1)(\tfrac12-2)\cdots(\tfrac12-n+1)}{n!}
  = \frac{\tfrac12 \cdot (-\tfrac12)\cdot(-\tfrac32)\cdots(\tfrac{3-2n}{2})}{n!}.
\]
There are $n$ factors of $2$ in the denominator, so a factor of $2^{-n}$ appears automatically. The remaining odd factors contribute a sign $(-1)^{n-1}$ and the product
\[
  1 \cdot 1 \cdot 3 \cdot 5 \cdots (2n-3) = (2n-3)!!.
\]
To convert the double factorial into an ordinary factorial, note that
\[
  (2m)! = (1 \cdot 3 \cdot 5 \cdots (2m-1))(2 \cdot 4 \cdot 6 \cdots 2m)
  = (2m-1)!! \cdot 2^m m!.
\]
Hence
\[
  (2m-1)!! = \frac{(2m)!}{2^m m!}.
\]
Setting $m=n-1$ gives
\[
  (2n-3)!! = \frac{(2n-2)!}{2^{n-1}(n-1)!}.
\]
Substituting this back into the formula for $\binom{1/2}{n}$, we obtain
\begin{align*}
  \binom{1/2}{n}
  &= \frac{(-1)^{n-1}}{2^n n!}(2n-3)!! \\
  &= \frac{(-1)^{n-1}}{2^n n!}\cdot \frac{(2n-2)!}{2^{n-1}(n-1)!} \\
  &= \frac{(-1)^{n-1}(2n-2)!}{2^{2n-1} n! \, (n-1)!} \\
  &= \frac{(-1)^{n-1}}{2^{2n-1}n}\binom{2n-2}{n-1}
  \qquad (n \ge 1).
\end{align*}

Now substitute the binomial expansion into $C(z)$:
\[
  C(z)
  = \frac{1}{2z}\Bigl(1-\sum_{n \ge 0}\binom{1/2}{n}(-4z)^n\Bigr)
  = -\frac{1}{2z}\sum_{n \ge 1}\binom{1/2}{n}(-4z)^n.
\]
The coefficient of $z^n$ is therefore
\[
  [z^n]C(z) = -\frac12 \binom{1/2}{n+1}(-4)^{n+1}.
\]
Using the formula above with $n+1$ in place of $n$,
\[
  [z^n]C(z)
  = -\frac12 \cdot \frac{(-1)^n}{2^{2n+1}(n+1)}\binom{2n}{n} \cdot (-1)^{n+1}4^{n+1}.
\]
Now
\[
  (-1)^n(-1)^{n+1} = -1,
\]
so the two minus signs cancel:
\[
  [z^n]C(z)
  = \frac{4^{n+1}}{2^{2n+2}(n+1)}\binom{2n}{n}
  = \frac{1}{n+1}\binom{2n}{n}.
\]
Thus
\[
  C_n = \frac{1}{n+1}\binom{2n}{n},
\]
the standard closed form for the $n$-th Catalan number.

\begin{remark}
The formula for $C(z)$ shows that this generating function is \emph{algebraic}: it satisfies a polynomial equation over $\mathbb{Q}(z)$. What we have \emph{not} proved yet is the full analytic story about its singularities and the asymptotic behavior of $C_n$. Chapter~5 will show that the square-root singularity at $z=1/4$ leads to
\[
  C_n \sim \frac{4^n}{n^{3/2}\sqrt{\pi}},
\]
and will state carefully the hypotheses under which an $n^{-3/2}$ factor really appears. Unambiguous context-free grammars form an important later example, but that connection is only a preview at this stage.
\end{remark}

\section*{Ordinary versus exponential generating functions}

The generating functions above are \emph{ordinary generating functions} (OGFs): the sequence element $a_n$ is literally the coefficient of $z^n$. For \emph{unlabeled} combinatorial structures, where we care only about size and shape, OGFs are the natural choice.

But many problems in combinatorics --- and later, in probabilistic models of strings --- involve \emph{labeled} structures, where the $n$ atoms in a structure carry distinct labels from $\{1,\dots,n\}$ and the assignment of labels matters. The canonical example is permutations: the number of permutations of $n$ distinct elements is $n!$. Combining two labeled structures requires choosing which labels go where. If an $A$-structure uses $k$ labels and a $B$-structure uses the remaining $n-k$, then one must first choose the $k$ labels for the $A$-part. For example, when $n=4$ and $k=2$, there are $\binom{4}{2}=6$ ways to decide which two labels go to the $A$-structure. In general the total count is
\[
  \sum_{k=0}^n \binom{n}{k} a_k b_{n-k}.
\]
This is called \emph{binomial convolution}. The extra binomial coefficient is exactly the cost of distributing labels.

\begin{definition}
The \emph{exponential generating function} (EGF) of a sequence $(a_n)_{n \ge 0}$ is
\[
  \hat{A}(z) \;=\; \sum_{n \ge 0} a_n \frac{z^n}{n!}.
\]
\end{definition}

The key algebraic fact is that binomial convolution corresponds to ordinary multiplication of EGFs. Multiply the series and collect all pairs of terms whose total degree is $n$:
\[
  [z^n](\hat{A}\cdot\hat{B})
  = \sum_{k=0}^n \frac{a_k}{k!}\frac{b_{n-k}}{(n-k)!}
  = \frac{1}{n!}\sum_{k=0}^n \binom{n}{k}a_k b_{n-k},
\]
since $\binom{n}{k}=n!/(k!(n-k)!)$. So
\[
  n!\,[z^n](\hat{A}\cdot\hat{B})
\]
is exactly the binomial convolution, which is why labeled structures prefer EGFs.

We write $[z^n/n!]\hat{A}(z)$ as shorthand for $n!\,[z^n]\hat{A}(z)=a_n$, that is, the coefficient of $z^n/n!$ in the EGF.

\begin{example}
Consider the sequence $a_n=n!$, counting permutations. Its OGF is
\[
  \sum_{n \ge 0} n! \, z^n.
\]
For any fixed $z \ne 0$, the ratio of successive terms has absolute value
\[
  \frac{(n+1)!|z|^{n+1}}{n!|z|^n} = (n+1)|z| \to \infty,
\]
so this series does not converge analytically anywhere except at $z=0$. Its EGF, on the other hand, is
\[
  \hat{S}(z)
  = \sum_{n \ge 0} n!\frac{z^n}{n!}
  = \sum_{n \ge 0} z^n
  = \frac{1}{1-z},
\]
by the geometric-series lemma. The coefficient $[z^n/n!]\hat{S}(z)=1$ means that there is one underlying \emph{linear-order shape} of size $n$; the $n!$ different labelings of that shape are precisely the $n!$ permutations.
\end{example}

The distinction between OGF and EGF is not merely aesthetic: using an OGF for a labeled structure or an EGF for an unlabeled one leads to incorrect functional equations. The \emph{symbolic method}, developed in Chapter~3, makes this assignment systematic. There we will define the constructions $\SEQ$, $\SET$, $\CYC$, and $\MSET$ and translate them into generating-function operations.

\section*{Road map}

This chapter has introduced generating functions as formal algebraic objects. The exact results proved here were algebraic: recurrences became functional equations, and those equations led to closed forms. The asymptotic remarks were only previews. The next chapter treats $z$ as a complex variable and develops the analytic tools needed to justify those previews. One reason complex numbers matter is that singularities away from the real axis can still control the coefficients of a power series. Chapter~3 then introduces the symbolic method, turning combinatorial grammar rules directly into generating-function equations. Chapters~4 and~5 return to the Fibonacci and Catalan examples and prove carefully how poles and square-root singularities control coefficient growth. Later, when we connect these ideas to language models, we will also need to distinguish carefully between counting generating functions and probability-weighted generating functions rather than treating them as the same object.

\begin{exercise}
Find the ordinary generating function for the sequence $a_n = 3^n$ (ternary strings of length $n$) by both methods: (a) summing the geometric series directly, and (b) applying the recurrence-to-GF recipe to the recurrence $a_n = 3a_{n-1}$, $a_0 = 1$.
\end{exercise}

\begin{exercise}
Use the generating function method to solve the recurrence $a_n = 2a_{n-1} + 1$ for $n \ge 1$, with $a_0 = 0$. (Hint: multiply by $z^n$ and sum. You will need to handle $\sum_{n \ge 1} z^n = z/(1-z)$ for the constant term on the right side. Use partial fractions on the resulting $A(z)$ to find a closed form for $a_n$.)
\end{exercise}

\begin{exercise}
Verify the generalized binomial coefficient formula $\binom{1/2}{n} = (-1)^{n-1}/(2^{2n-1} n) \cdot \binom{2n-2}{n-1}$ for $n = 1, 2, 3, 4$ by computing both sides directly.
\end{exercise}

\section*{Further Reading}

\begin{itemize}
  \item \textbf{P. Flajolet and R. Sedgewick}, \emph{Analytic Combinatorics} (Cambridge University Press, 2009) \cite{FlajoletSedgewick2009} --- The definitive reference for analytic combinatorics. Part~A (Chapters~I--II) covers formal power series, OGFs, EGFs, and the translation dictionary between combinatorial constructions and generating-function operations with full proofs and extensive examples.

  \item \textbf{A. Nijenhuis and H.\,S. Wilf}, \emph{Combinatorial Algorithms} (Academic Press, 2nd ed., 1978) \cite{NijenhuisWilf1978} --- Focuses on algorithms for random generation and enumeration of combinatorial structures; includes generating-function derivations as part of the algorithmic analysis, with a hands-on computational style that complements the more theoretical approach of the present chapter.

  \item \textbf{H.\,S. Wilf}, \emph{generatingfunctionology} (2nd ed., Academic Press, 1994; freely available online) --- A short, readable introduction to generating functions aimed at undergraduates. Covers OGFs and EGFs through worked problems; an excellent companion for anyone finding the present chapter's pace too quick.

  \item \textbf{R.\,P. Stanley}, \emph{Enumerative Combinatorics}, Vol.~1 (Cambridge, 1997) --- Chapter~1 gives a thorough algebraic treatment of formal power series and the ring $\mathbb{Q}[[z]]$; Chapter~4 covers rational and algebraic generating functions in depth appropriate for graduate study.

  \item \textbf{D.\,E. Knuth}, \emph{The Art of Computer Programming}, Vol.~1, \S1.2.9 --- A classic account of generating functions as a tool for solving recurrences, with worked examples drawn from algorithm analysis; especially useful for readers interested in the connection to computer science.
\end{itemize}
