\chapter{Just Enough Complex Analysis}

Chapter~1 treated $z$ as a formal symbol, manipulated according to algebraic rules with no thought of convergence or magnitude. The real power of analytic combinatorics comes from promoting $z$ to a genuine complex variable and studying how the coefficients $[z^n]A(z)$ are controlled by the analytic behavior of the resulting function $A(z)$. This chapter develops the minimum complex analysis needed for that move.

Some of the material below will be proved carefully: the root test, the Cauchy--Hadamard formula for the radius of convergence, the coefficient-extraction formula on a circle, and the basic exponential bound that follows from it. Other facts are genuinely deeper theorems of complex analysis. When that happens, I will say so explicitly and use them as black boxes rather than pretending they are obvious. The standard treatment of all topics below is \cite{FlajoletSedgewick2009}, Part~A.

\section{Complex Numbers and Convergence}

Recall that
\[
  \mathbb{C} = \{x+iy : x,y \in \mathbb{R}\}, \qquad i^2=-1.
\]
The \emph{modulus} of $z=x+iy$ is
\[
  |z| = \sqrt{x^2+y^2}.
\]
If $z \ne 0$, we can also describe it by an angle. Angles are only defined modulo $2\pi$, and here we choose the \emph{principal argument} $\arg z \in (-\pi,\pi]$.

Euler's formula
\[
  e^{i\theta} = \cos\theta + i\sin\theta
\]
is a standard identity, derivable from the power-series definitions of $e^x$, $\sin x$, and $\cos x$. It allows every nonzero complex number to be written in \emph{polar form}
\[
  z = r e^{i\theta},
\]
where $r=|z|>0$ and $\theta=\arg z$. Multiplication becomes especially transparent in this form:
\[
  r_1 e^{i\theta_1}\cdot r_2 e^{i\theta_2}
  = r_1 r_2 e^{i(\theta_1+\theta_2)},
\]
because exponentials add exponents. So moduli multiply and arguments add.

Convergence in $\mathbb{C}$ is measured by the modulus: a sequence $(z_k)$ converges to $L$ if $|z_k-L| \to 0$. A series $\sum_k w_k$ \emph{converges absolutely} if $\sum_k |w_k| < \infty$.

\begin{proposition}
If $\sum_k |w_k|$ converges, then $\sum_k w_k$ converges in $\mathbb{C}$.
\end{proposition}

\begin{proof}
Write $w_k = u_k + iv_k$ with $u_k,v_k \in \mathbb{R}$. Then
\[
  |u_k| \le |w_k|, \qquad |v_k| \le |w_k|.
\]
So $\sum_k u_k$ and $\sum_k v_k$ both converge absolutely as real series. Hence
\[
  \sum_k w_k = \sum_k u_k + i\sum_k v_k
\]
converges in $\mathbb{C}$.
\end{proof}

The geometric series is the first analytic example. For every integer $N \ge 0$,
\[
  \sum_{k=0}^{N} w^k = \frac{1-w^{N+1}}{1-w}, \qquad w \ne 1.
\]
If $|w|<1$, then $|w|^{N+1} \to 0$, so the finite sums converge to
\[
  \sum_{k \ge 0} w^k = \frac{1}{1-w}.
\]
This is the same identity as in ordinary real algebra; the only extra ingredient is that the modulus tells us exactly when $w^{N+1} \to 0$.

\section{Power Series and Radius of Convergence}

Let
\[
  A(z) = \sum_{n \ge 0} a_n z^n
\]
be a power series with coefficients in $\mathbb{C}$. Treating $z$ as a complex variable, the set of $z$ for which the series converges absolutely is always a disk centered at the origin. What happens on the boundary of that disk can be subtle: some series converge at certain boundary points and diverge at others.

\begin{definition}
For a real sequence $(x_n)$, the \emph{limit superior} $\limsup_{n \to \infty} x_n$ is the largest subsequential limit of the sequence. Equivalently, it is the eventual upper envelope of the tail values $x_N,x_{N+1},x_{N+2},\dots$ as $N \to \infty$.
\end{definition}

\begin{theorem}[Root test]
Let $\sum_{n \ge 0} u_n$ be a series of complex numbers, and let
\[
  \lambda = \limsup_{n \to \infty} |u_n|^{1/n}.
\]
If $\lambda < 1$, then the series converges absolutely. If $\lambda > 1$, then $u_n \not\to 0$, so the series diverges.
If $\lambda = 1$, the root test by itself does not decide what happens.
\end{theorem}

\begin{proof}
Suppose $\lambda < 1$. Choose a number $r$ with
\[
  \lambda < r < 1.
\]
By the definition of $\limsup$, all sufficiently large terms satisfy
\[
  |u_n|^{1/n} \le r.
\]
So for all large $n$,
\[
  |u_n| \le r^n.
\]
Since $\sum r^n$ is a convergent geometric series, comparison shows that $\sum |u_n|$ converges, so $\sum u_n$ converges absolutely.

Now suppose $\lambda > 1$. Choose $r$ with
\[
  1 < r < \lambda.
\]
Then infinitely many $n$ satisfy $|u_n|^{1/n} \ge r$; otherwise all sufficiently large terms would have $|u_n|^{1/n} < r$, forcing $\limsup |u_n|^{1/n} \le r$, a contradiction. For those infinitely many $n$,
\[
  |u_n| \ge r^n \ge 1.
\]
So $u_n$ cannot tend to $0$, and the series diverges.
\end{proof}

\begin{theorem}[Cauchy--Hadamard]
Let
\[
  A(z) = \sum_{n \ge 0} a_n z^n,
  \qquad
  L = \limsup_{n \to \infty} |a_n|^{1/n}.
\]
Define the \emph{radius of convergence} $R$ by
\[
  \frac{1}{R} = L,
\]
with the conventions $1/\infty = 0$ and $1/0 = \infty$. Then the series converges absolutely for every $z$ with $|z|<R$ and diverges for every $z$ with $|z|>R$.
\end{theorem}

\begin{proof}
Apply the root test to the series
\[
  \sum_{n \ge 0} a_n z^n.
\]
Its $n$th term is $u_n = a_n z^n$, so
\[
  |u_n|^{1/n} = |a_n|^{1/n}|z|.
\]
Therefore
\[
  \limsup_{n \to \infty} |u_n|^{1/n}
  = |z| \limsup_{n \to \infty} |a_n|^{1/n}
  = |z|L
  = \frac{|z|}{R}.
\]
If $|z|<R$, then $|z|/R < 1$, so the root test gives absolute convergence. If $|z|>R$, then $|z|/R > 1$, so the terms do not tend to $0$ and the series diverges.
\end{proof}

The extreme cases fit naturally into the same statement. If $R=\infty$, the series converges for every complex $z$; such a function is called \emph{entire}. If $R=0$, the series diverges for every nonzero $z$.

The combinatorial meaning of the radius is that it captures exponential growth. When $0<R<\infty$, define
\[
  \theta(n) = a_n R^n.
\]
Then
\[
  |\theta(n)|^{1/n} = |a_n|^{1/n}R,
\]
so
\[
  \limsup_{n \to \infty} |\theta(n)|^{1/n}
  = R \limsup_{n \to \infty} |a_n|^{1/n}
  = 1.
\]
Thus one may write
\[
  a_n = R^{-n}\theta(n),
\]
where the factor $\theta(n)$ is \emph{subexponential} in the sense above. Later chapters will show how the singularity type controls this finer factor.

\begin{example}
For
\[
  B(z) = \frac{1}{1-2z} = \sum_{n \ge 0} 2^n z^n,
\]
we have $|2^n|^{1/n}=2$, so $R=1/2$. The growth rate $2^n = (1/2)^{-n}$ matches the rule $R^{-n}$ exactly.

For the Fibonacci generating function
\[
  F(z) = \frac{z}{1-z-z^2},
\]
the denominator vanishes at
\[
  z = \frac{-1 \pm \sqrt{5}}{2}.
\]
These are approximately $0.618$ and $-1.618$, so the root of smaller modulus is $1/\varphi \approx 0.618$, where $\varphi=(1+\sqrt{5})/2$. Thus the nearest singularity lies at distance $1/\varphi$ from the origin, so $R=1/\varphi$. The later asymptotic statement
\[
  F_n \sim \frac{\varphi^n}{\sqrt{5}}
\]
is a preview of the transfer machinery proved in Chapter~4, not something earned from the radius alone.
\end{example}

\section{Analytic Functions}

\begin{definition}
An \emph{open set} $U \subseteq \mathbb{C}$ is a region such that every point of $U$ is surrounded by a small disk lying completely inside $U$.
\end{definition}

\begin{definition}
A function $f:U \to \mathbb{C}$ is \emph{holomorphic} on $U$ (also called \emph{analytic}) if it is complex-differentiable at every point of $U$.
\end{definition}

\begin{theorem}[Basic equivalence, used without proof]
For a function on an open set, the following conditions are equivalent:
\begin{enumerate}
\item it is complex-differentiable;
\item it satisfies the Cauchy--Riemann equations and is infinitely differentiable;
\item it is locally equal to a convergent power series.
\end{enumerate}
\end{theorem}

This equivalence is one of the deepest beginner-level theorems in complex analysis. We will use it as a black box. The surprising part is that complex differentiability is much more rigid than real differentiability: a real differentiable function need not equal any power series, but a holomorphic function automatically does.

\begin{theorem}[Power-series calculus, used without proof]
Inside its disk of convergence, a power series may be differentiated and integrated term by term, and the resulting power series has the same radius of convergence.
\end{theorem}

As a consequence, if
\[
  A(z) = \sum_{n \ge 0} a_n z^n
\]
has radius of convergence $R>0$, then it defines a holomorphic function on the disk $|z|<R$.

\begin{definition}
Let $f$ be holomorphic on a punctured disk
\[
  0 < |z-z_0| < \varepsilon.
\]
If $f$ cannot be extended holomorphically to the center point $z_0$, then $z_0$ is called a \emph{singularity} of $f$.
Any singularity of minimal modulus is called \emph{dominant}; there may be several.
\end{definition}

\begin{theorem}[Nearest-singularity principle, used without proof]
The radius of convergence of a power series centered at $0$ equals the distance from $0$ to the nearest singularity of the holomorphic function represented by the series.
\end{theorem}

This theorem is one of the main bridges from complex analysis to asymptotics. It explains why complex singularities matter even when the coefficients are real and positive.

\section{The Geometric Series as Prototype}

The function
\[
  f(z) = \frac{1}{1-z}
\]
is the model example. For $|z|<1$ the geometric series converges absolutely, and its sum is
\[
  \frac{1}{1-z} = \sum_{n \ge 0} z^n.
\]
The equality is valid only inside the disk $|z|<1$ when interpreted analytically.

\begin{definition}
A function has a \emph{simple pole} at $z=\rho$ if $(z-\rho)f(z)$ stays finite and nonzero as $z \to \rho$.
\end{definition}

The function $1/(1-z)$ is holomorphic on $\mathbb{C}\setminus\{1\}$ and has a simple pole at $z=1$. The same formula still makes sense at points such as $z=2$ or $z=-3$, even though the power series $\sum z^n$ diverges there. Extending a holomorphic function beyond its original disk of convergence in this way is called \emph{analytic continuation}. The uniqueness of analytic continuation is another major theorem that we will not prove here.

This example already illustrates the nearest-singularity principle: the singularity at $z=1$ sits exactly one unit from the origin, so the radius of convergence is $1$.

\section{Recovering Coefficients from a Circle}

To extract coefficients analytically, we integrate around circles. If $g$ is continuous on $|z|=r$, we define
\[
  \oint_{|z|=r} g(z)\,dz
\]
to mean the ordinary real integral obtained by parameterizing the circle as
\[
  z = re^{it}, \qquad 0 \le t \le 2\pi,
\]
so that
\[
  dz = i r e^{it}\,dt.
\]

\begin{proposition}[Coefficient formula on a circle]
Let
\[
  f(z) = \sum_{m \ge 0} a_m z^m
\]
have radius of convergence $R$, and choose $0<r<R$. Then for every $n \ge 0$,
\[
  a_n = [z^n]f(z) = \frac{1}{2\pi i}\oint_{|z|=r} \frac{f(z)}{z^{n+1}}\,dz.
\]
\end{proposition}

\begin{proof}
Because $r<R$, the series for $f(re^{it})$ converges absolutely and uniformly in $t$, so we may integrate term by term. Parameterizing the circle gives
\[
  \frac{1}{2\pi i}\oint_{|z|=r} \frac{f(z)}{z^{n+1}}\,dz
  = \frac{1}{2\pi}\int_0^{2\pi} f(re^{it})\, r^{-n} e^{-int}\,dt.
\]
Now expand
\[
  f(re^{it}) = \sum_{m \ge 0} a_m r^m e^{imt}.
\]
Substituting,
\[
  \frac{1}{2\pi}\int_0^{2\pi} \sum_{m \ge 0} a_m r^{m-n} e^{i(m-n)t}\,dt
  = \sum_{m \ge 0} a_m r^{m-n} \cdot \frac{1}{2\pi}\int_0^{2\pi} e^{i(m-n)t}\,dt.
\]
The integral on the right is $1$ when $m=n$ and $0$ otherwise, so only one term survives:
\[
  \frac{1}{2\pi i}\oint_{|z|=r} \frac{f(z)}{z^{n+1}}\,dz = a_n.
\]
\end{proof}

This is the most concrete version of coefficient extraction. A deeper theorem, Cauchy's integral formula, extends the same idea to arbitrary holomorphic functions and derivatives; we will use that broader result later as a black box.

For estimates we also need the size of $f$ on the circle. Because $f$ is continuous and the circle $|z|=r$ is compact, the maximum
\[
  M(r) = \max_{|z|=r} |f(z)|
\]
exists.

\begin{proposition}[Circle bound]
With the same hypotheses as above,
\[
  |a_n| \le \frac{M(r)}{r^n}.
\]
\end{proposition}

\begin{proof}
Starting from the coefficient formula and parameterizing the circle,
\[
  |a_n|
  \le \frac{1}{2\pi}\int_0^{2\pi} |f(re^{it})|\, r^{-n}\,dt
  \le \frac{1}{2\pi}\int_0^{2\pi} M(r)\, r^{-n}\,dt
  = \frac{M(r)}{r^n}.
\]
\end{proof}

This gives an exponential upper bound. If we can choose a larger circle, the factor $r^{-n}$ improves. A deeper theorem of complex analysis says that, as long as we stay in a region where $f$ is holomorphic, such contour integrals may be deformed without changing their value. We will use that theorem as a black box. The important caution is this: the bound above explains why nearby singularities control the \emph{best exponential rate}, but it does \emph{not} by itself determine the exact asymptotic formula. For that, one needs local analysis near the singularity.

\section{Prototype Singularities}

This section is a catalogue of model cases, not the full transfer theorem. We compute the coefficients for a few prototype singular behaviors and use them as guides. Chapter~4 will state the general theorem that turns these prototypes into systematic asymptotics.

For generating functions with nonnegative coefficients, a classical theorem of Pringsheim says that the point $z=R$ on the positive real axis is always a singularity. We will use that theorem as a black box when needed. There may still be additional dominant singularities of the same modulus elsewhere in the complex plane.

Before discussing square roots and logarithms, one warning is needed. In complex analysis, $\sqrt{\cdot}$ and $\log$ are not globally single-valued functions. To make sense of them analytically, one chooses a \emph{branch}. Near the origin we use the principal branch, which is holomorphic on a slit domain obtained by removing a ray from the singular point. Since our power-series expansions are centered at $0$, this local branch is the one that matters.

\textbf{Poles.}
For a positive integer $k$,
\[
  f(z) = \frac{1}{(1-z/\rho)^k}
\]
has a pole of order $k$ at $z=\rho$. Using the generalized binomial theorem from Chapter~1 with $\alpha=-k$ and $w=-z/\rho$, one obtains
\[
  \frac{1}{(1-z/\rho)^k}
  = \sum_{n \ge 0} \binom{n+k-1}{k-1}\rho^{-n}z^n.
\]
Thus
\[
  [z^n]f(z) = \binom{n+k-1}{k-1}\rho^{-n}.
\]
The binomial coefficient is a polynomial in $n$ of degree $k-1$. Indeed,
\[
  \binom{n+k-1}{k-1}
  = \frac{(n+k-1)(n+k-2)\cdots(n+1)}{(k-1)!},
\]
so dividing by $n^{k-1}$ gives
\[
  \frac{\binom{n+k-1}{k-1}}{n^{k-1}}
  = \frac{(1+\frac{k-1}{n})(1+\frac{k-2}{n})\cdots(1+\frac{1}{n})}{(k-1)!}
  \to \frac{1}{(k-1)!}.
\]
Hence
\[
  \binom{n+k-1}{k-1} \sim \frac{n^{k-1}}{(k-1)!}.
\]
So a simple pole contributes a pure exponential $\rho^{-n}$, a double pole contributes about $n\rho^{-n}$, a triple pole contributes about $\frac{n^2}{2}\rho^{-n}$, and so on.

\textbf{Square-root branch points.}
Consider the prototype
\[
  f(z) = \sqrt{1-z/\rho} = (1-z/\rho)^{1/2}.
\]
By the generalized binomial theorem,
\[
  f(z) = \sum_{n \ge 0} \binom{1/2}{n}(-z/\rho)^n.
\]
Chapter~1 proved that
\[
  \binom{1/2}{n} = \frac{(-1)^{n-1}}{2^{2n-1}n}\binom{2n-2}{n-1}
  \qquad (n \ge 1),
\]
so for $n \ge 1$,
\[
  [z^n]f(z)
  = -\frac{1}{2^{2n-1}n}\binom{2n-2}{n-1}\rho^{-n}.
\]
Now apply Stirling's theorem, which we use here as a standard asymptotic fact:
\[
  m! \sim \sqrt{2\pi m}\Bigl(\frac{m}{e}\Bigr)^m.
\]
With $m=n-1$,
\[
  \binom{2m}{m}
  = \frac{(2m)!}{(m!)^2}
  \sim \frac{\sqrt{4\pi m}(2m/e)^{2m}}{2\pi m (m/e)^{2m}}
  = \frac{4^m}{\sqrt{\pi m}}.
\]
Therefore
\[
  \binom{2n-2}{n-1} \sim \frac{4^{n-1}}{\sqrt{\pi n}},
\]
and hence
\[
  [z^n]\sqrt{1-z/\rho}
  \sim -\frac{1}{2\sqrt{\pi}}\, n^{-3/2}\rho^{-n}.
\]
This is a \emph{prototype} calculation: it suggests the $n^{-3/2}$ behavior that later chapters will prove for a broad class of genuine square-root singularities.

\textbf{Logarithmic singularities.}
Inside $|w|<1$, the geometric series may be integrated term by term:
\[
  \frac{1}{1-w} = \sum_{n \ge 0} w^n.
\]
Integrating from $0$ to $w$ gives
\[
  -\log(1-w) = \sum_{n \ge 1} \frac{w^n}{n},
\]
where we use the branch of the logarithm that is real at $w=0$. Substituting $w=z/\rho$ yields
\[
  \log\frac{1}{1-z/\rho} = \sum_{n \ge 1} \frac{\rho^{-n}}{n} z^n,
\]
so
\[
  [z^n]\log\frac{1}{1-z/\rho} = \frac{\rho^{-n}}{n}.
\]
The correction factor is now $1/n$. Logarithms arise naturally in labelled structures involving cycles; Chapter~3 will define the symbolic operators $\SET$ and $\CYC$ that lead to such terms.

\textbf{Essential singularities.}
Functions such as
\[
  e^{1/(1-z)}
\]
have an essential singularity at $z=1$. We mention the term only for orientation: its precise definition uses Laurent series, which we do not develop here. The point is that its behavior near the singularity is much more irregular than the clean pole and branch-point cases above, so the elementary catalogue is no longer enough.

\section{The Big Picture}

The moral of this chapter is simple, but it must be stated carefully.

\begin{enumerate}
\item The radius of convergence tells us the exponential scale of the coefficients.
\item Nearby singularities explain why that exponential scale is what it is.
\item Model singularities suggest what the subexponential correction should look like.
\end{enumerate}

What we do \emph{not} yet have is the general theorem turning the third step into a rigorous automatic procedure. That theorem is the transfer theorem of Flajolet and Odlyzko \cite{FlajoletOdlyzko1990}, developed in Chapter~4 and explained in detail in \cite{FlajoletSedgewick2009}.

So the practical program is:
\begin{enumerate}
\item derive a usable generating function by algebra or combinatorial structure;
\item locate the dominant singularities on the boundary of the disk of convergence;
\item identify the local singular behavior near those points;
\item only then apply an appropriate transfer theorem.
\end{enumerate}

Even this summary comes with caveats: there may be several dominant singularities, branch choices may matter, and upper bounds alone do not yield exact asymptotics. Still, the basic philosophy is already visible: singularities are not annoyances to be avoided but the source of coefficient information.

\begin{exercise}
Find the radius of convergence of
\[
  \sum_{n \ge 1} \frac{z^n}{n^2}.
\]
The series also converges at $z=1$. Explain why convergence at a boundary point does not settle what happens to analytic continuation beyond the disk $|z|<1$.
\end{exercise}

\begin{exercise}
Show that
\[
  \frac{1}{(1-z)^3} = \sum_{n \ge 0} \binom{n+2}{2} z^n
\]
by applying the generalized binomial theorem from Chapter~1 with $\alpha=-3$ and $w=-z$. Then verify that
\[
  \binom{n+2}{2} \sim \frac{n^2}{2},
\]
consistent with the pole catalogue above.
\end{exercise}

\begin{exercise}
The series
\[
  \sum_{n \ge 0} n!\,z^n
\]
has radius of convergence $R=0$; verify this using Cauchy--Hadamard. Why is this an example where the formal power series from Chapter~1 is still a perfectly valid algebraic object, but does not define an analytic function on any nontrivial disk around $0$?
\end{exercise}

\section*{Further Reading}

\begin{itemize}
  \item \textbf{P. Flajolet and R. Sedgewick}, \emph{Analytic Combinatorics} (Cambridge University Press, 2009) \cite{FlajoletSedgewick2009}, Part~B --- Chapters~IV--VI develop the complex-analytic foundations used throughout this book: analytic continuation, the transfer theorem, and singularity analysis in full generality. Reading Part~B alongside the present chapter is strongly recommended.

  \item \textbf{P. Flajolet and A.\,M. Odlyzko}, ``Singularity analysis of generating functions,'' \emph{SIAM J.\ Discrete Math.} (1990) \cite{FlajoletOdlyzko1990} --- The original paper introducing the transfer theorem and $\Delta$-domains; more concise than the textbook treatment and useful for seeing the theorem in its original form.

  \item \textbf{L.\,V. Ahlfors}, \emph{Complex Analysis} (3rd ed., McGraw-Hill, 1979) --- The standard graduate-level reference for classical complex analysis. Chapters~1--4 cover all the material in the present chapter (convergence, analyticity, Cauchy's theorem, singularity classification) with full proofs; Chapter~5 introduces residues and contour integration.

  \item \textbf{E.\,C. Titchmarsh}, \emph{The Theory of Functions} (2nd ed., Oxford, 1939) --- A more classical treatment, particularly strong on power series (Chapter~1) and contour integration and residues (Chapter~3); useful as a second reference for the Cauchy--Hadamard theorem and the classification of isolated singularities.

  \item \textbf{P. Henrici}, \emph{Applied and Computational Complex Analysis}, Vol.~1 (Wiley, 1974) --- Covers power series and analytic functions with an emphasis on effective computation and coefficient extraction; bridges the gap between pure complex analysis and the algorithmic side of generating functions.
\end{itemize}
