\chapter{Weighted Finite Automata and Rational Power Series}
\label{ch07:top}

Chapter~6 established finite automata as devices that assign a binary verdict---accept or reject---to every string over a finite alphabet. The present chapter replaces that verdict with a weight drawn from an algebraic structure called a semiring. Each string $w$ receives a weight $f(w) \in K$, and the resulting object is a \emph{weighted finite automaton} (WFA). When $K = \mathbb{R}_{\ge 0}$ and the weights are normalized to sum to one, the WFA becomes a probabilistic generative model over $\Sigma^*$. The central theorem of the chapter is that the length-indexed generating function of any such model is automatically a rational function of $z$, bringing the full machinery of Part~I---partial fractions, dominant-pole asymptotics, entropy rates via spectral radii---to bear on probabilistic language models. This rationality result is not an approximation or a limiting statement; it is an exact algebraic consequence of the matrix structure of WFAs, and it is the engine driving the WFA-approximation program developed in Chapter~12.

\section{Semirings}

A \emph{semiring} $(K, \oplus, \otimes, \mathbf{0}, \mathbf{1})$ is a set $K$ equipped with two binary operations satisfying the following axioms: $(K, \oplus, \mathbf{0})$ is a commutative monoid; $(K, \otimes, \mathbf{1})$ is a monoid; $\otimes$ distributes over $\oplus$ from both sides; and $\mathbf{0} \otimes a = a \otimes \mathbf{0} = \mathbf{0}$ for all $a \in K$. The crucial omission relative to a ring is that $\oplus$ need not have inverses. The name ``semi'' reflects exactly this missing structure.

Two semirings dominate the present chapter. The \emph{Boolean semiring} $(\{0,1\}, \vee, \wedge, 0, 1)$ reduces the theory to classical unweighted automata: transition weights are either present or absent, matrix multiplication becomes the Boolean matrix product, and the WFA accepts a string iff the computed weight is $1$. The \emph{nonneg\-ative real semiring} $(\mathbb{R}_{\ge 0}, +, \times, 0, 1)$ produces the probabilistic theory. One may also work over $\mathbb{R}$ or $\mathbb{C}$ (discarding the nonnegativity constraint), over the tropical semiring $(\mathbb{R} \cup \{+\infty\}, \min, +, +\infty, 0)$ for optimization problems, or over the formal power series ring $\mathbb{R}[[z]]$ itself. The general theory of WFAs over arbitrary semirings is developed in \cite{DrosteKuichVogler2009}; for present purposes the nonnegative reals suffice.

\section{Weighted Finite Automata}

\begin{definition}
A \emph{weighted finite automaton} (WFA) over alphabet $\Sigma$ with $n$ states and weights in semiring $K$ is a triple $\mathcal{A} = (\alpha, \{A_\sigma\}_{\sigma \in \Sigma}, \beta)$ where:
\begin{itemize}
  \item $\alpha \in K^{1 \times n}$ is a row vector of initial weights,
  \item $\beta \in K^{n \times 1}$ is a column vector of final weights,
  \item $A_\sigma \in K^{n \times n}$ is a transition matrix for each letter $\sigma \in \Sigma$.
\end{itemize}
The weight assigned to a string $w = \sigma_1 \sigma_2 \cdots \sigma_k$ is
\begin{equation}\label{eq:wfa-weight}
  f_\mathcal{A}(w) = \alpha \cdot A_{\sigma_1} A_{\sigma_2} \cdots A_{\sigma_k} \cdot \beta \in K.
\end{equation}
The empty string $\varepsilon$ receives weight $f_\mathcal{A}(\varepsilon) = \alpha \cdot \beta$.
\end{definition}

The computation \eqref{eq:wfa-weight} has a concrete state-sum interpretation. Define the weight of a path $q_0 \xrightarrow{\sigma_1} q_1 \xrightarrow{\sigma_2} \cdots \xrightarrow{\sigma_k} q_k$ as $\alpha_{q_0} \cdot (A_{\sigma_1})_{q_0, q_1} \cdots (A_{\sigma_k})_{q_{k-1}, q_k} \cdot \beta_{q_k}$. Then $f_\mathcal{A}(w)$ is the $\oplus$-sum of this quantity over all paths labeled by $w$, which is precisely the $(1,1)$ entry of the matrix product in \eqref{eq:wfa-weight} once one accounts for the vector boundary conditions. In the Boolean semiring this collapses to: $f_\mathcal{A}(w) = 1$ iff at least one accepting path exists, recovering the classical Myhill-Nerode picture from Chapter~6.

\begin{example}[Fibonacci WFA]\label{ex:fibonacci-wfa}
Let $\Sigma = \{a\}$, $n = 2$, $K = \mathbb{R}$. Set
\[
  \alpha = \begin{pmatrix} 1 & 0 \end{pmatrix}, \quad
  \beta = \begin{pmatrix} 0 \\ 1 \end{pmatrix}, \quad
  A_a = \begin{pmatrix} 0 & 1 \\ 1 & 1 \end{pmatrix}.
\]
Then $f_\mathcal{A}(a^k) = \alpha A_a^k \beta$. One verifies directly:
$k=1$: $\alpha A_a \beta = (1,0)\begin{pmatrix}0&1\\1&1\end{pmatrix}\begin{pmatrix}0\\1\end{pmatrix} = (1,0)\begin{pmatrix}1\\1\end{pmatrix} = 1 = F_1$.
$k=2$: $\alpha A_a^2 \beta = (1,0)\begin{pmatrix}1&1\\1&2\end{pmatrix}\begin{pmatrix}0\\1\end{pmatrix} = (1,0)\begin{pmatrix}1\\2\end{pmatrix} = 1 = F_2$.
$k=3$: $A_a^3 = A_a^2 \cdot A_a = \begin{pmatrix}1&1\\1&2\end{pmatrix}\begin{pmatrix}0&1\\1&1\end{pmatrix} = \begin{pmatrix}1&2\\2&3\end{pmatrix}$, so $\alpha A_a^3 \beta = (1,0)\begin{pmatrix}1&2\\2&3\end{pmatrix}\begin{pmatrix}0\\1\end{pmatrix} = 2 = F_3$.
In general $f_\mathcal{A}(a^k) = F_k$, the $k$-th Fibonacci number (with $F_0 = 0$, $F_1 = 1$). The two-state automaton encodes the linear recurrence $F_k = F_{k-1} + F_{k-2}$ in the companion matrix $A_a$.
\end{example}

\section{Stochastic WFAs and Probabilistic Language Models}

Specialize now to $K = \mathbb{R}_{\ge 0}$. A WFA is \emph{stochastic} if $\alpha$ is a probability distribution over states (i.e., $\sum_q \alpha_q = 1$, $\alpha_q \ge 0$) and if the following row-stochasticity condition holds: for each state $q$,
\[
  \sum_{\sigma \in \Sigma} \sum_{q'} (A_\sigma)_{q,q'} + \beta_q = 1.
\]
The interpretation is that from state $q$, the automaton either emits a letter $\sigma$ and transitions to state $q'$ (with probability $(A_\sigma)_{q,q'}$) or emits the end-of-sequence symbol $\EOS$ and halts (with probability $\beta_q$). Under this condition $f_\mathcal{A}(w)$ is exactly the probability that the automaton generates the string $w$ and then halts, and the weights sum to one over all $w \in \Sigma^*$.

Stochastic WFAs generalize \emph{hidden Markov models} (HMMs). In a standard HMM there is a Markov chain over hidden states and at each step a symbol is emitted from a state-dependent emission distribution; one may always write such a model as a stochastic WFA by absorbing the emission distributions into the transition matrices $A_\sigma$. The converse does not hold without further structural conditions: a stochastic WFA can encode transitions that depend simultaneously on both the current and next state in ways that are not directly factored as emission times transition. The representational difference is analytically subtle but matters for expressiveness arguments in Section~6.

\section{The Generating Function is Rational}

Fix a WFA $\mathcal{A} = (\alpha, \{A_\sigma\}, \beta)$ over $K = \mathbb{R}$ (or $\mathbb{C}$). Define the \emph{aggregate transition matrix}
\[
  A := \sum_{\sigma \in \Sigma} A_\sigma \in \mathbb{R}^{n \times n}.
\]
The length-indexed generating function of $\mathcal{A}$ is
\[
  F(z) := \sum_{w \in \Sigma^*} f_\mathcal{A}(w)\, z^{|w|}.
\]
Group terms by length:
\[
  F(z) = \sum_{k \ge 0} z^k \!\!\!\sum_{\substack{w \in \Sigma^* \\ |w| = k}} f_\mathcal{A}(w)
       = \sum_{k \ge 0} z^k \!\!\!\sum_{\substack{\sigma_1,\ldots,\sigma_k \\ \in \Sigma}} \alpha A_{\sigma_1} \cdots A_{\sigma_k} \beta.
\]
The key step: the multilinear expansion of a product of sums gives
\[
  \sum_{\sigma_1,\ldots,\sigma_k \in \Sigma} A_{\sigma_1} \cdots A_{\sigma_k} = \Bigl(\sum_{\sigma \in \Sigma} A_\sigma\Bigr)^k = A^k,
\]
because the $A_\sigma$ are matrices over a ring (not merely a semiring), so ordinary matrix algebra applies. Therefore
\[
  F(z) = \sum_{k \ge 0} z^k\, \alpha A^k \beta = \alpha \Bigl(\sum_{k \ge 0} (zA)^k\Bigr) \beta.
\]
The matrix power series $\sum_{k \ge 0} (zA)^k$ converges to $(I - zA)^{-1}$ whenever $|z| < 1/\rho(A)$, where $\rho(A)$ denotes the spectral radius of $A$. This yields:

\begin{theorem}[Rationality of WFA Generating Functions]\label{thm:wfa-rational}
For any WFA $\mathcal{A} = (\alpha, \{A_\sigma\}, \beta)$ over $\mathbb{R}$ (or $\mathbb{C}$), the length-indexed generating function is rational:
\[
  F(z) = \alpha (I - zA)^{-1} \beta,
\]
where $A = \sum_{\sigma \in \Sigma} A_\sigma$. This rational function has poles only at the reciprocals of the nonzero eigenvalues of $A$, and the power series converges for $|z| < 1/\rho(A)$.
\end{theorem}

The numerator polynomial arises from Cramer's rule applied to $(I - zA)^{-1}$: each entry is a ratio of degree-$(n-1)$ and degree-$n$ polynomials in $z$, so $F(z) = P(z)/Q(z)$ with $\deg P \le n-1$ and $Q(z) = \det(I - zA)$ the characteristic polynomial of $A$ (with variable substitution $\lambda \mapsto 1/z$).

The connection to Part~I is immediate. Chapter~4 established that the coefficients of a rational generating function $P(z)/Q(z)$ satisfy a linear recurrence with characteristic polynomial $Q$, and the asymptotic behavior is governed by the pole of smallest modulus (i.e., the reciprocal of the dominant eigenvalue). Precisely, if $\lambda_1$ is the eigenvalue of $A$ with largest absolute value and multiplicity $m$, then
\[
  [z^n] F(z) \sim C \lambda_1^n n^{m-1}
\]
for some constant $C$ depending on the eigenvector structure and the boundary vectors $\alpha$, $\beta$.

For stochastic WFAs the row-stochasticity condition $A\mathbf{1} + \beta = \mathbf{1}$ with $\beta \ge 0$ implies $A\mathbf{1} \le \mathbf{1}$, so $A$ is substochastic and has spectral radius $\rho(A) \le 1$. When the automaton defines a proper probability distribution---a condition analyzed in depth in Chapter~8---one has $\rho(A) < 1$, the generating function $F(z)$ has no pole on the unit circle, and the string probabilities decay geometrically in length.

\begin{example}[Fibonacci asymptotics via Theorem~\ref{thm:wfa-rational}]
Returning to Example~\ref{ex:fibonacci-wfa}, the aggregate matrix is $A = A_a = \begin{pmatrix}0&1\\1&1\end{pmatrix}$. Its characteristic polynomial is $\lambda^2 - \lambda - 1 = 0$, with roots $\varphi = (1+\sqrt{5})/2$ (the golden ratio) and $\hat\varphi = (1-\sqrt{5})/2$. The generating function is
\[
  F(z) = \alpha(I - zA)^{-1}\beta = \frac{z}{1 - z - z^2},
\]
with poles at $z = 1/\varphi$ and $z = 1/\hat\varphi$. The dominant pole is $1/\varphi$, and partial fractions (exactly as in Chapter~1) give $[z^n]F(z) = F_n \sim \varphi^n/\sqrt{5}$. Theorem~\ref{thm:wfa-rational} recovers this from the WFA structure without any ad hoc computation.
\end{example}

\section{The Kleene--Schützenberger Theorem}

The classical Kleene theorem characterizes the languages accepted by finite automata as exactly the regular languages---those definable by regular expressions built from union, concatenation, and Kleene star. Schützenberger's generalization lifts this to the weighted setting. A formal power series $S = \sum_{w \in \Sigma^*} c_w \cdot w$ over a semiring $K$ is called \emph{recognizable} if it equals $f_\mathcal{A}$ for some WFA $\mathcal{A}$, and \emph{rational} if it belongs to the smallest family of series containing all polynomials (finite sums $\sum_{w} c_w \cdot w$) and closed under sum, product, and the operation $S^* := \sum_{k \ge 0} S^k$ (defined when the constant term of $S$ is $\mathbf{0}$).

\begin{theorem}[Kleene--Schützenberger, informal]
A formal power series $S \in K\langle\!\langle \Sigma^* \rangle\!\rangle$ is recognizable if and only if it is rational.
\end{theorem}

This result, proved in full generality in \cite{BerstelReutenauer2011}, is the noncommutative analogue of Kleene's theorem. The ``noncommutative'' qualifier is essential: the series lives over the free monoid $\Sigma^*$ where word concatenation is not commutative, so $S^*$ must be interpreted with care and the star is not the geometric series in a scalar variable but rather the fixed-point equation $S^* = \varepsilon + S \cdot S^*$. The connection to Theorem~\ref{thm:wfa-rational} is that when one commutes all the variables---replacing each word $w$ by its length $z^{|w|}$---the rational series maps to a rational function of $z$, recovering the scalar generating function.

\section{The Icard Hierarchy and What WFAs Cannot Express}

Theorem~\ref{thm:wfa-rational} implies a sharp limitation: the generating function of any WFA-definable distribution is rational, hence its Taylor coefficients satisfy a linear recurrence. But Chapter~5 introduced distributions whose generating functions are genuinely algebraic---satisfying a polynomial equation $P(z, F(z)) = 0$ but not a linear equation $Q(z) F(z) = R(z)$---and whose coefficients obey the $n^{-3/2}$ subexponential law characteristic of algebraic singularities. No WFA can reproduce such behavior.

This observation is formalized by the \emph{probabilistic Chomsky hierarchy} of \cite{Icard2020}. At the finite-state level sit the \emph{rational stochastic languages}: probability distributions over $\Sigma^*$ computable by some stochastic WFA. At the context-free level sit the \emph{algebraic stochastic languages}: distributions definable by probabilistic context-free grammars (PCFGs). Icard's non-collapse theorem states that these two classes are strictly distinct: there exist PCFG-definable distributions whose generating functions are algebraic but not rational, and no stochastic WFA matches them. The proof proceeds by exhibiting a PCFG for balanced-parentheses strings, whose length-distribution generating function satisfies $F = z + F^2$ (a quadratic, not a rational function), and arguing that no WFA can replicate this distribution.

Complementarily, \cite{SmithJohnson2007} showed that at the context-free level, adding arbitrary positive weights to a CFG (producing a general weighted CFG) does not increase the class of conditional parse distributions beyond what a properly normalized PCFG can represent. This means the additional flexibility of arbitrary weights buys nothing at the context-free level---the representational power is captured by the probabilistic normalization alone.

Together, these results carve out the following picture. WFAs occupy a well-defined and analytically tractable niche: their distributions are exactly the rational-stochastic languages, their generating functions are rational, and the full arsenal of Part~I applies. PCFGs occupy the algebraic level above, and Chapters~9--11 will develop their theory. The generating functions there satisfy polynomial equations, the Flajolet--Odlyzko transfer theorems for algebraic singularities apply, and the asymptotic coefficients reflect the tree-counting structure of context-free derivations. The general reference for the WFA theory is \cite{DrosteKuichVogler2009}; the generating function perspective is detailed in \cite{FlajoletSedgewick2009}.

\section{Significance for Language Models}

A large language model is neither a WFA nor a PCFG. It is a high-dimensional, nonlinear function---typically a transformer with billions of parameters---and its associated distribution over $\Sigma^*$ has no obvious algebraic characterization. Nevertheless, the WFA framework provides two concrete analytical tools.

First, if an LLM's next-token distribution can be \emph{approximated} by a WFA---in the sense that there exists a stochastic WFA $\mathcal{A}$ such that $\|f_{\mathcal{A}} - f_{\mathrm{LLM}}\|$ is small in some appropriate norm---then the LLM's generating function is approximately rational, and the Flajolet--Sedgewick machinery yields approximate asymptotics. The quality of the approximation is controlled by the number of states $n$: larger $n$ allows more complex rational approximations. Chapter~12 develops this WFA-approximation approach and gives conditions under which the approximation error vanishes as $n \to \infty$.

Second, the Icard hierarchy provides sharp obstructions. If empirical measurements reveal that a language model's length distribution follows a $n^{-3/2}$ law (as Chapter~5 suggests for certain grammatical structures), this is a signature of algebraic, not rational, behavior. No WFA approximation---regardless of the number of states---can reproduce this law exactly, because all rational functions have exponential or polynomial-times-exponential coefficient asymptotics, never a pure power law with negative fractional exponent. The algebraic structure must be modeled at the PCFG level or above. This is the content of Icard's non-collapse theorem applied to empirical data, and it provides a principled test for whether a finite-state approximation is adequate for a given modeling task.

The generating function $F(z) = \alpha(I - zA)^{-1}\beta$ of Theorem~\ref{thm:wfa-rational} will reappear throughout Part~III. It encodes the entropy rate of the WFA as $h = -\log \rho(A)$ (when $\mathcal{A}$ is stochastic and irreducible), connects to the spectral theory of Markov chains, and provides the baseline against which the richer algebraic behavior of PCFGs is measured. Chapter~8 addresses the foundational question that must precede all asymptotic analysis: when does a probabilistic grammar define a \emph{proper} probability distribution---one that assigns total mass one to $\Sigma^*$---and what algebraic conditions govern this consistency?

\section*{Further Reading}

\begin{itemize}
  \item \cite{DrosteKuichVogler2009} --- the comprehensive reference for weighted automata over arbitrary semirings, covering the algebraic foundations, closure properties, and decision problems; the standard source for the general theory presented in this chapter.
  \item \cite{BerstelReutenauer2011} --- develops rational series over free monoids from first principles, proving the Kleene--Schützenberger theorem in full generality and connecting recognizable series to matrix representations.
  \item \cite{Icard2020} --- establishes the probabilistic Chomsky hierarchy and the non-collapse theorem separating rational stochastic languages from algebraic ones; the direct source for the Icard hierarchy discussion in Section~6.
  \item \cite{SmithJohnson2007} --- shows that weighted CFGs with arbitrary positive weights are representationally equivalent to properly normalized PCFGs at the context-free level, settling a question about the expressiveness of unnormalized weighting schemes.
  \item \cite{FlajoletSedgewick2009} --- Chapter~V on applications of rational and meromorphic asymptotics provides the analytic-combinatorics counterpart to the WFA rationality theorem, including the transfer theorems that translate pole locations into coefficient asymptotics.
\end{itemize}
