\chapter{Autoregressive Language Models}
\label{ch10:top}

So far we have treated ``distributions over strings'' abstractly: a probability measure $\mu$ on $\Sigma^*$, a weighted automaton, a formal power series with nonnegative coefficients summing to one. This chapter introduces the concrete object that the rest of the book analyzes: a large language model. The exposition is deliberately stripped down. An LLM, for our purposes, is a probability distribution on $\Sigma^*$, nothing more. The neural network is the parameter, not the object. What matters is what distribution it induces, and what mathematical tools are available for analyzing that distribution. Once we have this, the singularity analysis of Part I and the entropy theory of Part II can be brought to bear directly.

\section{Tokens, Vocabulary, and End-of-Sequence}

An LLM operates over a finite vocabulary $\Sigma$, typically consisting of \emph{tokens} obtained by applying a subword encoding to a large text corpus. The three standard algorithms are byte-pair encoding (BPE), WordPiece, and the unigram language model. Each algorithm merges or segments character sequences according to corpus statistics, producing units that are neither words nor characters but lie somewhere between. Modern vocabularies range from $|\Sigma| \approx 30{,}000$ (as in GPT-2 and the Llama family) to $|\Sigma| \approx 200{,}000$ (as in GPT-4). The vocabulary always includes a distinguished \emph{end-of-sequence token} $\EOS \in \Sigma$. Every sampled string ends with the first occurrence of $\EOS$, at which point sampling halts. Strings not containing $\EOS$ in their interior are the only grammatical outputs; $\EOS$ appearing mid-string would be treated as a terminal event, truncating everything after it.

Tokenization is not a detail to be abstracted away. It determines what counts as an atom for the combinatorial class of LLM outputs. A byte-pair tokenizer running on English text will produce a granularity different from word-level and different from character-level. Zipf exponents on token frequencies, length distributions measured in tokens, and entropy rates all depend on which tokenizer was applied. Two models with identical weights but different tokenizers define genuinely different distributions on $\Sigma^*$. Part~V returns to this in Chapter~16, where tokenization artifacts appear as poles in the length generating function.

\section{The Autoregressive Factorization}

An LLM defines, for every history $h \in \Sigma^*$, a conditional distribution $p(\cdot \mid h) \in \Delta(\Sigma)$ over the next token, treating $\EOS$ as an ordinary token. Given these conditionals, the probability of generating a specific finite string $w = w_1 w_2 \cdots w_n$ followed immediately by $\EOS$ is
\[
  p(w \cdot \EOS) \;=\; p(\EOS \mid w_1 \cdots w_n) \prod_{t=1}^{n} p(w_t \mid w_1 \cdots w_{t-1}),
\]
where the empty product at $t = 0$ is $1$ (so the probability of the string consisting of $\EOS$ alone is $p(\EOS \mid \varepsilon)$). Sampling proceeds left to right: draw $w_1 \sim p(\cdot \mid \varepsilon)$, then $w_2 \sim p(\cdot \mid w_1)$, and so on, until $\EOS$ is sampled. This procedure is called \emph{autoregressive decoding}.

\begin{definition}[Autoregressive language model]
An \emph{autoregressive language model} over $\Sigma$ is a family $\mathcal{M} = \{p(\cdot \mid h)\}_{h \in \Sigma^*}$ of conditional distributions $p(\cdot \mid h) \in \Delta(\Sigma)$, one for each finite history $h \in \Sigma^*$. The induced distribution on $\Sigma^*$ is defined by the autoregressive factorization above.
\end{definition}

The parameters of the neural network are merely a compact representation of the family $\mathcal{M}$; the mathematical object is the family itself. Two distinct parameter vectors that induce the same family of conditionals define the same language model in our sense. This is the appropriate level of abstraction for the analytic combinatorics program: we care about generating functions of the output distribution, and those are determined by $\mathcal{M}$, not by the network's internal weight matrices.

\section{Softmax and the Role of Temperature}

In every modern LLM, the conditional $p(\cdot \mid h)$ is computed via \emph{softmax} applied to a vector of real-valued \emph{logits} $\ell(h) \in \mathbb{R}^{|\Sigma|}$:
\[
  p(\sigma \mid h) \;=\; \frac{e^{\ell_\sigma(h)}}{\displaystyle\sum_{\sigma' \in \Sigma} e^{\ell_{\sigma'}(h)}}.
\]
Softmax maps any real vector to a strictly positive probability distribution: every token, including every token that might look semantically bizarre in context, receives nonzero probability. This seemingly minor property has structural consequences that will dominate the next two sections.

\emph{Temperature scaling.} In practice one often samples from the tempered distribution
\[
  p_T(\sigma \mid h) \;=\; \frac{e^{\ell_\sigma(h)/T}}{\displaystyle\sum_{\sigma'} e^{\ell_{\sigma'}(h)/T}},
\]
where $T > 0$ is the \emph{sampling temperature}. At $T = 1$ one recovers the model's native distribution. As $T \to 0$, all mass concentrates on the $\operatorname{argmax}$ token (greedy decoding). As $T \to \infty$, $p_T(\cdot \mid h)$ converges to the uniform distribution on $\Sigma$. Temperature scaling is a one-parameter deformation of the model's output distribution; it preserves the support (every token remains reachable for any $T > 0$) while reshaping the entropy. Temperature is the central object of Part~IV, where its effect on the dominant singularity of the length PGF is analyzed.

\section{Transformers in One Paragraph}

A transformer maps the history $h$ to the logits $\ell(h)$ through a stack of $L$ layers, each consisting of a multi-head self-attention sublayer followed by a position-wise feedforward network, with residual connections and layer normalization at each sublayer \cite{VaswaniEtAl2017}. The formal object we care about is the input--output function $h \mapsto \ell(h)$; the internal mechanism is parameterization, not the subject of study. What matters analytically is that this function is (i) computable in polynomial time per step (quadratic in the context length via the attention computation), (ii) differentiable in the parameters, and (iii) structurally forced to produce strictly positive softmax outputs at every step. Property~(iii) is what the next section exploits. Nothing about the specific form of self-attention, feedforward width, or normalization scheme enters the tightness argument; the argument requires only that the logit function $h \mapsto \ell(h)$ maps into a bounded region of $\mathbb{R}^{|\Sigma|}$ for fixed parameters.

\section{Tightness for Transformer Language Models}

Chapter~8 defined \emph{tightness}: the autoregressive factorization yields a proper probability measure on $\Sigma^*$ (i.e., $\sum_{w \in \Sigma^*} p(w \cdot \EOS) = 1$) if and only if the sum $\sum_{t=1}^\infty \widetilde{p}_\EOS(t)$ diverges, where $\widetilde{p}_\EOS(t)$ is the probability that $\EOS$ has not yet been emitted by step $t$. For transformer language models, tightness is automatic by a result of Du et al.\ \cite{DuEtAl2023}.

The argument is essentially the following. The logit vector $\ell(h) \in \mathbb{R}^{|\Sigma|}$ depends on the history $h$ through the transformer, but because the transformer's parameters are a fixed finite collection of weight matrices and biases, and because attention weights and layer normalizations are bounded per step, there is a uniform constant $C > 0$ depending only on the parameters such that $|\ell_\sigma(h)| \le C$ for all $h \in \Sigma^*$ and all $\sigma \in \Sigma$. Then
\[
  p(\EOS \mid h) \;=\; \frac{e^{\ell_\EOS(h)}}{\displaystyle\sum_{\sigma \in \Sigma} e^{\ell_\sigma(h)}} \;\ge\; \frac{e^{-C}}{|\Sigma|\, e^{C}} \;=\; \frac{e^{-2C}}{|\Sigma|} \;=:\; \varepsilon \;>\; 0.
\]
The EOS probability is bounded below by a uniform constant $\varepsilon$ that depends only on the parameter bound $C$ and the vocabulary size. By Proposition~4.3 of Du et al., a uniform lower bound $p(\EOS \mid h) \ge \varepsilon > 0$ for all histories implies tightness: the probability of still generating at time $t$ is at most $(1 - \varepsilon)^t$, so $\widetilde{p}_\EOS(t) \le (1-\varepsilon)^t$, and the survival sum is bounded by a convergent geometric series, meaning generation terminates almost surely.

\begin{theorem}[Du et al.\ 2023, \cite{DuEtAl2023}]
Every softmax-output transformer language model with bounded parameters is tight: it defines a proper probability measure on $\Sigma^*$, assigning zero total mass to the set of infinite strings $\Sigma^\omega$.
\end{theorem}

This theorem is not a curiosity. It is the permission slip for everything that follows. All of the singularity-analysis machinery of Part~I presupposes that the object being analyzed is actually a probability distribution, not a defective measure leaking mass to $\Sigma^\omega$. Tightness is what guarantees that the length PGF $P(z) = \sum_{n \ge 0} p_n z^n$ satisfies $P(1) = 1$ and that the counting generating function of the typical set has a well-defined radius of convergence whose reciprocal encodes the entropy rate. Transformers come with this guarantee built in, for free, by the structure of softmax.

\begin{remark}
The tightness bound $\varepsilon = e^{-2C}/|\Sigma|$ is extremely loose for any well-trained model. In practice $p(\EOS \mid h)$ varies enormously with $h$, approaching $1$ at natural sentence boundaries and near $0$ mid-token. But for the purpose of the existence theorem, the uniform floor suffices.
\end{remark}

Contrast with non-tight recurrent models. Du et al.\ show by explicit construction that an RNN with ReLU activations can be non-tight: one can tune the network so that the EOS logit decays faster than any polynomial in the sequence length, causing $\sum_t \widetilde{p}_\EOS(t)$ to converge (their Example~2.4). Non-tightness is therefore a genuine pathology, not an artifact of poor training. It is avoidable only by architectural choice. The softmax output layer is exactly such a choice; other choices, including hard gates or ReLU activations applied directly to the EOS logit, do not yield the same guarantee. This makes the softmax operation load-bearing in a mathematical sense that goes beyond its role as a differentiable normalization.

\section{Generating Functions for a Concrete LLM}

Chapters~8 and~9 distinguished two generating functions associated with a string-valued distribution. For a concrete LLM both are well-defined objects, and the distinction matters.

\emph{Length PGF.} Define $p_n = \Pr_{\mathcal{M}}(|W| = n)$, the probability that the model generates a string of exactly $n$ tokens before emitting $\EOS$ (not counting the $\EOS$ token itself). The length probability generating function is
\[
  P(z) \;=\; \sum_{n \ge 0} p_n z^n.
\]
For a tight model, $P(1) = 1$. The singularity of $P$ closest to the origin on the positive real axis --- whether it lies at $z = 1$, beyond $z = 1$, or is a branch point --- controls how the tail $p_n$ decays as $n \to \infty$. This is the natural object when asking: ``how long does the LLM generate before emitting $\EOS$?'' A model with a simple pole of $P$ at $z_0 > 1$ produces geometrically decaying length tails; a model whose $P$ has a branch point at $z = 1$ produces polynomial tails.

\emph{Counting GF.} For $\alpha \in (0,1)$, define the \emph{$\alpha$-typical set} $T_n^\alpha$ to be the set of strings $w$ of length $n$ with $p(w) \ge 2^{-n(h+\alpha)}$, where $h$ is the entropy rate of $\mathcal{M}$. The counting generating function of the typical set is
\[
  A(z) \;=\; \sum_{n \ge 0} |T_n^\alpha| \, z^n.
\]
By the asymptotic equipartition property, $|T_n^\alpha| \sim 2^{nh}$ for large $n$, so the radius of convergence of $A(z)$ is $R = 2^{-h}$, giving $h = \log_2(1/R)$. This is the natural object when asking: ``how large is the set of likely outputs at length $n$?'' The dominant singularity of $A$ at $z = R$ connects the entropy rate of the model to the analytic structure of the counting series.

The two objects address two different questions, and the literature on each is largely disjoint. The length PGF $P$ belongs to the domain of stopping-time analysis; the counting GF $A$ belongs to the domain of information-theoretic combinatorics. Part~V makes the separation explicit through the ``Gap~1 vs.\ Gap~2'' dichotomy \cite{CotterellEtAl2023}: Gap~1 concerns the discrepancy between a model's probability and a reference measure at fixed length (related to $A$), while Gap~2 concerns the probability assigned to strings beyond a certain length (related to $P$). When someone says ``the generating function of an LLM,'' one must always ask which of these is meant.

\section{An LLM as a Very-High-Order Markov Source}

The autoregressive factorization makes an LLM a Markov chain on the state space $\Sigma^*$ (the entire set of finite prefixes) with transition kernel $p(\cdot \mid h)$. This is an enormously high-dimensional, indeed uncountably branching, Markov chain, but it is still a Markov chain in the formal sense, and the theory of entropy rate, stationary distributions, and mixing from Part~II Chapter~9 applies formally. The entropy rate
\[
  h(\mathcal{M}) \;=\; \lim_{n \to \infty} \frac{1}{n} H(W_1, \ldots, W_n)
\]
exists for any tight autoregressive model by the subadditivity of entropy.

What is structurally new in the LLM setting is the scale of state dependence. A classical $k$-th order Markov source has transition kernel depending only on the last $k$ tokens; its entropy rate can be estimated from $(k+1)$-gram statistics, and its generating functions are rational by the transfer matrix theorem of Chapters~6--7. An LLM with full self-attention, by contrast, has transition kernel $p(\cdot \mid h)$ depending on the entire prefix $h$: the distribution over the next token can depend, in principle, on the token at position one. The long-range attention patterns break the finite-Markov-order structure entirely.

This has a direct consequence for the singularity-analysis program. The transfer matrix theorem guarantees rational GFs for finite-state Markov sources. If an LLM's attention machinery lets it approximate context-free or even more exotic behavior (Chapter~11), then the natural question becomes: does the LLM's output distribution lie in the rational, algebraic, or some wilder class of formal power series, and what does the dominant singularity of its length PGF or counting GF look like? These are precisely the questions of Parts~IV and~V.

\begin{example}
A unigram model, $p(\sigma \mid h) = q(\sigma)$ for a fixed distribution $q$, is a zero-order Markov source. Its length PGF is $P(z) = q(\EOS) z / (1 - (1 - q(\EOS))z)$, a rational function with a simple pole at $z = 1/(1 - q(\EOS)) > 1$. Its counting GF has radius $2^{-H(q \setminus \EOS)}$ where $H(q \setminus \EOS)$ is the per-token entropy over non-EOS tokens. A full transformer deforms both of these objects in ways that depend on the corpus and the architecture, and the analytic structure of the resulting GFs is the subject of the rest of this book.
\end{example}

Chapter~11 maps the transformer architecture against the formal language hierarchy, and makes precise why vanilla transformers without chain-of-thought are constrained to operate near the rational-stochastic corner of that hierarchy. This matters because it implies that the output GFs of such models are expected to be ``close to rational''---they should have a small number of dominant poles, with subdominant algebraic or essential singularities suppressed---and therefore amenable to the weighted finite automaton approximation program developed in Chapter~12. The tightness theorem of this chapter is the necessary prerequisite: it tells us the object whose singularities we are about to study is a genuine probability measure, not a ghost.
