\chapter{Motif Statistics and Analytic Information Theory}
\label{ch17}

This chapter is the most classical in Part~V. The questions we ask --- how often does a pattern appear in a random string? how deep is a random trie? what is the typical information content of a long sequence? --- were answered, for Markov sources, in the 1990s and early 2000s by a school of researchers centered on Wojciech Szpankowski. The toolkit they developed is called \emph{analytic information theory}, and it is precisely the toolkit the LLM community should reach for when it wants rigorous answers about sequence statistics.

\section{What Is Analytic Information Theory?}
\label{ch17:ait}

Analytic information theory is the research program of applying complex-analytic methods --- Mellin transforms, saddle-point integration, Cauchy's coefficient formula, singularity analysis --- to classical problems in information theory: entropy rates, data compression, pattern matching, suffix trees, tries. The program is associated above all with the book of Szpankowski~\cite{Szpankowski2001}, which remains the definitive reference.

\begin{remark}
Szpankowski's book is organized by \emph{technique}, not by problem: one chapter for poissonization, one for the Mellin transform, one for the saddle-point method, one for singularity analysis (in the sense of Flajolet--Sedgewick~\cite{FlajoletSedgewick2009}). This is deliberate. The same complex-analytic machine applies to problems that look unrelated at the problem level. A reader who learns the techniques acquires a reusable kit.
\end{remark}

The core observation is that many combinatorial and probabilistic quantities --- expected values, variances, limit distributions --- can be read off from the analytic behavior of a generating function near its dominant singularities. For sequences generated by a Markov source (or, more generally, a finite-state automaton), the relevant generating functions are rational, which makes the singularity analysis particularly clean.

The LLM connection is direct: as argued in Chapter~12, an LLM approximated by a WFA is essentially a high-dimensional Markov source. Every result in analytic information theory that holds for Markov sources therefore applies, in principle, to WFA-approximated LLMs. The remainder of this chapter makes this precise.

\section{Motif Occurrence Generating Functions}
\label{ch17:motif-gf}

Let $\Sigma$ be a finite alphabet of size $\sigma$, and let $M$ be a \emph{motif} --- either a fixed word or, more generally, the language recognized by some regular expression. We want to study, in a random string over $\Sigma$, the number of occurrences of~$M$.

\begin{definition}[Bivariate motif generating function]
\label{ch17:def-bivariate-gf}
The \emph{bivariate generating function} for strings counted by length and motif count is
\[
  F(z, u) = \sum_{w \in \Sigma^*} z^{|w|}\, u^{\#M(w)},
\]
where $\#M(w)$ denotes the number of (possibly overlapping) occurrences of $M$ in $w$, and each string $w$ is weighted by the probability assigned to it by the source.
\end{definition}

For a \emph{uniform} source over $\Sigma$, every word of length $n$ has weight $\sigma^{-n}$, so we set $z \mapsto z/\sigma$ and count unweighted. For a Markov source the weights are products of transition probabilities.

The coefficient $[z^n u^k] F(z,u)$ then gives the (weighted) number of strings of length $n$ containing exactly $k$ occurrences of $M$. Extracting asymptotics amounts to analyzing $F(z, u)$ for $u$ near $1$ and $z$ near its dominant singularity.

\subsection{The Autocorrelation Polynomial}

The structure of $F(z, u)$ is controlled by the \emph{autocorrelation polynomial} of $M$, introduced by Guibas and Odlyzko~\cite{GuibasOdlyzko1981}.

\begin{definition}[Autocorrelation polynomial]
\label{ch17:def-autocorrelation}
Let $M = m_1 m_2 \cdots m_\ell$ be a word of length $\ell$. The autocorrelation polynomial is
\[
  C(z) = \sum_{k=0}^{\ell-1} c_k\, z^k,
\]
where $c_k = 1$ if the suffix of $M$ of length $k$ equals the prefix of $M$ of length $k$, and $c_k = 0$ otherwise. Equivalently, $c_k = [m_{\ell-k+1}\cdots m_\ell = m_1 \cdots m_k]$.
\end{definition}

The autocorrelation polynomial encodes how $M$ can overlap with itself: $c_k = 1$ means $M$ can begin at position $\ell - k + 1$ of a previous occurrence of $M$ and still produce a valid occurrence. This self-overlap structure determines, via the theory of Guibas--Odlyzko, the generating function for the waiting time until the first occurrence of $M$ in a random string, and thence the bivariate GF.

For random texts modeled by a Markov source, and for $M$ a word or a regular expression, the general bivariate-GF framework is developed in Nicodème--Salvy--Flajolet~\cite{NicodemeSalvyFlajolet2002}. The recipe is:

\begin{enumerate}
  \item Build a DFA $\mathcal{A}_{k}$ recognizing the language of strings containing \emph{exactly $k$ occurrences} of $M$ (or, more efficiently, a single DFA tracking the current occurrence count).
  \item Form the \emph{transfer matrix} $T(z, u)$ of the weighted automaton, where a transition reading letter $a$ from source state $i$ to state $j$ contributes a factor of $z \cdot p_{ij}$ (the source transition probability) and $u$ if a new occurrence of $M$ is completed.
  \item The bivariate GF is the rational function $F(z, u) = \mathbf{v}^T (I - T(z,u))^{-1} \mathbf{e}$ for appropriate initial vector $\mathbf{v}$ and final-state indicator $\mathbf{e}$.
  \item Singularity analysis: for fixed $u$ near $1$, the dominant singularity $z_0(u)$ of $F(z, u)$ is the zero of $\det(I - T(z,u))$ of smallest modulus. Perturbing $u$ from $1$ perturbs $z_0(u)$, and the derivatives of $z_0(u)$ at $u = 1$ give the mean and variance of $\#M$ in a string of length~$n$.
\end{enumerate}

This is entirely mechanical once the automaton is specified. It is one of the cleanest applications of the symbolic method and transfer-matrix technology developed in Chapters~4--6.

\begin{proposition}[Gaussian limit law for motif counts; Nicodème--Salvy--Flajolet~\cite{NicodemeSalvyFlajolet2002}]
\label{ch17:prop-gaussian}
Let $M$ be a motif recognized by a finite automaton, and let the source be an ergodic Markov chain on $\Sigma$. Let $X_n$ denote the number of occurrences of $M$ in a random string of length $n$. Then:
\[
  \mathbb{E}\, X_n = \mu n + O(1), \qquad \mathrm{Var}\, X_n = \sigma^2 n + O(1),
\]
for constants $\mu > 0$ and $\sigma^2 > 0$ computable from the transfer matrix, and
\[
  \frac{X_n - \mu n}{\sigma \sqrt{n}} \xrightarrow{d} \mathcal{N}(0, 1).
\]
\end{proposition}

The proof uses singularity perturbation: one shows that $\log F(z, e^t)$ has a simple pole at $z_0(e^t)$ whose location is analytic in $t$ near $0$, and the cumulant generating function of $X_n$ then satisfies the hypotheses of the quasi-power theorem (see Flajolet--Sedgewick~\cite{FlajoletSedgewick2009}, Chapter~IX). The Gaussian limit follows.

\section{Worked Example: The Motif ``aba''}
\label{ch17:example-aba}

\begin{example}[Autocorrelation of ``aba'']
\label{ch17:ex-aba}
Let $\Sigma = \{a, b\}$ with uniform probabilities $p_a = p_b = 1/2$, and let $M = \texttt{aba}$, a word of length $\ell = 3$.

We compute the autocorrelation polynomial $C(z) = c_0 + c_1 z + c_2 z^2$.

\begin{itemize}
  \item $c_0 = 1$ always (the empty overlap).
  \item $c_1$: is the suffix of $M$ of length $1$, namely $\texttt{a}$, equal to the prefix of length $1$, namely $\texttt{a}$? Yes. So $c_1 = 1$.
  \item $c_2$: is the suffix of $M$ of length $2$, namely $\texttt{ba}$, equal to the prefix of length $2$, namely $\texttt{ab}$? No. So $c_2 = 0$.
\end{itemize}

Therefore $C(z) = 1 + z$.

This means \texttt{aba} can overlap with a subsequent occurrence of itself only at offset $2$ (one character back), because the single-character suffix \texttt{a} of \texttt{aba} matches the single-character prefix \texttt{a}. Concretely, in the string \texttt{ababa}, the substring \texttt{aba} occurs at positions $1$ and $3$, overlapping in the \texttt{a} at position $3$.

The bivariate generating function for strings over $\{a,b\}$ with uniform weights has the form, in the Guibas--Odlyzko--Nicodème framework:
\[
  F(z, u) = \frac{C(z/2)}{C(z/2) - (u-1) z^\ell / 2^\ell}
            \cdot \frac{1}{1 - z},
\]
where the correction from the autocorrelation polynomial modifies the pole location.

More precisely, the expected number of occurrences of \texttt{aba} in a random string of length $n$ is
\[
  \mathbb{E}\, X_n = \frac{n - 2}{8} + O(1) \;\approx\; \frac{n}{8},
\]
since $\mu = p_a^2 p_b = (1/2)^2 \cdot (1/2) = 1/8$. The autocorrelation affects the variance and the higher-order terms but not the leading mean.
\end{example}

\section{Trie Depth and Markov Source Analysis}
\label{ch17:trie}

A \emph{trie} (retrieval tree) built from $n$ strings $w_1, \ldots, w_n \in \Sigma^\omega$ is the rooted tree in which each string corresponds to a root-to-leaf path, and internal nodes branch on the next character. The \emph{depth} $D_n$ of a random string in the trie --- equivalently, the length of the longest common prefix of a random string with all others --- is a fundamental quantity: it measures how long one must read a string before it becomes distinguishable.

Jacquet and Szpankowski~\cite{JacquetSzpankowski1991} analyzed trie depth under an ergodic Markov source:

\begin{proposition}[Trie depth under Markov source; Jacquet--Szpankowski~\cite{JacquetSzpankowski1991}]
\label{ch17:trie-depth}
Let strings be generated by an ergodic Markov chain on $\Sigma$ with entropy rate $h_1$ (in nats, or bits if logarithms are base $2$). Then the expected depth of a random key in a trie built from $n$ independent strings satisfies
\[
  \mathbb{E}\, D_n \sim \frac{1}{h_1} \log n \quad \text{as } n \to \infty.
\]
More precisely, there is a periodic fluctuation: $\mathbb{E}\, D_n = \frac{1}{h_1} \log n + \Lambda(\log n) + O(1/n)$, where $\Lambda$ is a small-amplitude periodic function coming from poles of a Mellin transform off the real axis.
\end{proposition}

The constant $1/h_1$ is the reciprocal of the entropy rate. For the binary uniform source, $h_1 = \log 2$ and $\mathbb{E}\, D_n \sim \log_2 n$. For $n = 1000$, this gives $\mathbb{E}\, D_{1000} \approx \log_2 1000 \approx 9.97 \approx 10$. A random binary string requires about $10$ bits before it is distinguished from $1000$ competitors --- exactly what information theory predicts, now with a proof via analytic methods.

The interpretation for language modeling is immediate. If we model a natural-language corpus as generated by a Markov source (or a WFA approximation of an LLM), then $D_n \sim \frac{1}{h}\log n$ says that the typical \emph{distinguishing context length} grows logarithmically in the corpus size, with the entropy rate as the determining constant.

\section{Connection to Chapter 9: The Entropy Rate Revisited}
\label{ch17:entropy-connection}

In Chapter~9, we established the identity
\[
  h = \log\frac{1}{R},
\]
where $R$ is the radius of convergence of the generating function counting strings by probability weight, and $h$ is the (Shannon) entropy rate of the source. The dominant singularity of the counting GF is at $z = R$, and $h$ is determined by its location.

In the language of Perron--Frobenius theory: for a primitive (irreducible, aperiodic) non-negative matrix $T$, the spectral radius $\rho(T)$ is the unique largest eigenvalue, with positive eigenvectors. For a Markov chain with transition matrix $P$, the relevant matrix in the trie-depth and motif-statistics problems is a \emph{weighted} version of $P$, with weights chosen to count string length. The dominant singularity of the transfer matrix GF occurs at $z = 1/\rho(T_z)$, where $T_z$ is $P$ weighted by $z$.

The entropy rate satisfies
\[
  h = -\sum_{i,j} \pi_i P_{ij} \log P_{ij},
\]
where $\pi$ is the stationary distribution. This is exactly $\log(1/R)$ when $R$ is the radius of convergence of the path-counting GF. The Perron--Frobenius theorem guarantees that this dominant singularity is a square-root or simple pole, not an accumulation point, which is why singularity analysis applies cleanly.

The trie-depth constant $1/h_1$ in Proposition~\ref{ch17:trie-depth} is the same $h$ appearing in Chapter~9, computed by the same spectral method. The two analyses are not independent: they are two applications of the same underlying complex-analytic framework to the same spectral radius.

\section{The Saddle-Point Method and Mellin Transforms}
\label{ch17:methods}

Szpankowski's book is organized around two complex-analytic tools that go beyond the algebraic singularity analysis of Flajolet--Sedgewick~\cite{FlajoletSedgewick2009}. We sketch them briefly.

\subsection{The Saddle-Point Method}
\label{ch17:saddle-point}

Singularity analysis extracts $[z^n] f(z)$ when $f$ has an algebraic or logarithmic singularity at a finite point $z_0$. But some generating functions are entire --- no finite singularity at all --- and grow like $e^{g(z)}$ for some $g$. The prototypical example is $e^z$, with $[z^n] e^z = 1/n!$.

For such functions, one uses the \emph{saddle-point method}. By Cauchy's formula,
\[
  [z^n] f(z) = \frac{1}{2\pi i} \oint \frac{f(z)}{z^{n+1}}\, dz.
\]
The integrand $h(z) = f(z) / z^{n+1}$ has a saddle point at $z_0$ where $h'(z_0) = 0$, equivalently where $f'(z_0)/f(z_0) = (n+1)/z_0$. One deforms the contour to pass through $z_0$ in the direction of steepest descent, so the integrand is concentrated near $z_0$ and the remaining integral is Gaussian. The result is
\[
  [z^n] f(z) \approx \frac{f(z_0)}{z_0^n \sqrt{2\pi \kappa_2}},
\]
where $\kappa_2$ is the second cumulant of the local Gaussian. Hayman's theorem gives sufficient conditions for this approximation to be valid for entire functions of so-called \emph{admissible} type.

In analytic information theory, the saddle-point method appears when one analyzes the distribution of the number of distinct words in a random string, or the profile of a random trie (how many nodes are at depth $k$). These quantities involve entire GFs and require saddle-point analysis for their asymptotics.

\subsection{The Mellin Transform}
\label{ch17:mellin}

The Mellin transform of a function $F: (0,\infty) \to \mathbb{R}$ is
\[
  \mathcal{M}[F](s) = \int_0^\infty x^{s-1} F(x)\, dx,
\]
defined for $s$ in a vertical strip of the complex plane. The Mellin transform is particularly suited to functions of the form $F(x) = \sum_k f(a_k x)$, which arise naturally when one counts strings by combining independent contributions from each position or branch.

The key property: if $F(x) = \sum_{k \geq 1} f(k x)$, then
\[
  \mathcal{M}[F](s) = \mathcal{M}[f](s) \cdot \zeta(s),
\]
where $\zeta$ is the Riemann zeta function (or its analog for the relevant Dirichlet series). The poles of $\mathcal{M}[F](s)$ in the complex plane then correspond to terms in the asymptotic expansion of $F(x)$ via Mellin--Perron inversion.

In trie analysis, the Mellin transform handles the sum over depths $k$ of the contribution of strings reaching depth $k$. The poles of the Mellin transform at non-real values of $s$ produce the periodic fluctuation $\Lambda(\log n)$ in Proposition~\ref{ch17:trie-depth}.

\section{Why This Matters for LLMs}
\label{ch17:llm-connection}

The results of this chapter were proved for Markov sources. Chapter~12 argued that LLMs can be approximated by WFAs, and WFAs are precisely weighted finite automata --- a generalization of Markov chains in which the weights need not be non-negative or normalized, but which still admit a transfer-matrix representation.

For a WFA approximation $A$ of an LLM:

\begin{itemize}
  \item The \emph{motif-count} analysis of Section~\ref{ch17:motif-gf} applies directly. Build the product automaton of $A$ with the motif-tracking DFA; extract the bivariate GF; apply singularity perturbation. This gives the mean, variance, and Gaussian limit law for the number of occurrences of any fixed phrase in LLM-generated text.

  \item The \emph{trie-depth} result of Section~\ref{ch17:trie} applies once the effective entropy rate $h_A$ of the WFA is defined. For an ergodic WFA, $h_A = \log(1/R_A)$ where $R_A$ is the spectral radius of the weighted transition matrix, exactly as in Chapter~9.

  \item The \emph{typical-set} analysis --- what is the set of strings an LLM assigns high probability to? --- is equivalent to a trie-depth or entropy-rate computation, and is therefore solved by the same methods.
\end{itemize}

\begin{remark}
There is a practical caveat. The WFA approximation of a large Transformer is high-dimensional: the transfer matrix may have dimension in the thousands or millions. The spectral quantities are then expensive to compute, though not conceptually different. This is a computational, not a mathematical, obstacle.
\end{remark}

The open problem, which we call Gap~9 and return to in Chapter~18, is this: \emph{no one has yet carried out the full Szpankowski-style analysis for a WFA approximation of a production LLM}. The mathematical pipeline exists. The WFA approximation methods of Chapter~12 exist. What is missing is the engineering to connect them, and perhaps the WFA approximations of sufficient quality to make the results meaningful.

This is not a fundamental obstacle. It is a gap between two mature bodies of work that have not yet been combined.

\section{Notes and Further Reading}
\label{ch17:notes}

The foundational reference for analytic information theory is Szpankowski~\cite{Szpankowski2001}. The autocorrelation polynomial and the exact GF for waiting times until a pattern first appears are due to Guibas and Odlyzko~\cite{GuibasOdlyzko1981}; their paper remains a model of elegant combinatorial argument. The extension to regular-expression motifs, including the bivariate GF and the Gaussian limit law stated in Proposition~\ref{ch17:prop-gaussian}, is in Nicodème--Salvy--Flajolet~\cite{NicodemeSalvyFlajolet2002}.

Trie depth under Markov sources was analyzed by Jacquet and Szpankowski~\cite{JacquetSzpankowski1991}; the same authors have a long series of papers on digital search trees, suffix trees, and Lempel--Ziv compression under Markov sources, all applying the Mellin-transform and saddle-point machinery. The Flajolet--Sedgewick book~\cite{FlajoletSedgewick2009} covers singularity analysis (Chapter~VI) and the saddle-point method (Chapter~VIII) systematically.

For readers who want to understand the periodic fluctuations in trie depth and digital tree statistics, the key tool is the Mellin--Perron formula combined with Lindstr\"om--Gessel--Viennot for lattice paths; Szpankowski~\cite{Szpankowski2001} Chapter~8 is the place to start.

The Hayman admissibility criterion for the saddle-point method is in Hayman's 1956 paper; see also Flajolet--Sedgewick~\cite{FlajoletSedgewick2009} Section~VIII.5 for a modern treatment.

\section*{Exercises}

\begin{exercise}
\label{ch17:ex-autocorrelation-abab}
Compute the autocorrelation polynomial for the motif $M = \texttt{abab}$ over $\Sigma = \{a, b\}$. Identify all non-trivial overlaps (those with $c_k = 1$ for $k \geq 1$), and explain in terms of the string structure why each overlap exists or does not exist. How does having $c_2 = 1$ affect the expected waiting time until the first occurrence of $M$ in a uniform random string, compared to a motif of the same length with $C(z) = 1$?
\end{exercise}

\begin{exercise}
\label{ch17:ex-entropy-trie}
Let $\Sigma = \{0, 1\}$ and consider the Markov chain with transition matrix
\[
  P = \begin{pmatrix} 1 - \alpha & \alpha \\ \beta & 1 - \beta \end{pmatrix}
\]
for parameters $\alpha, \beta \in (0, 1)$.
\begin{enumerate}
  \item[(a)] Compute the stationary distribution $\pi$.
  \item[(b)] Compute the entropy rate $h_1 = -\sum_{i,j} \pi_i P_{ij} \log P_{ij}$ (in bits).
  \item[(c)] For $n = 10^6$ strings generated by this chain, estimate the expected trie depth $\mathbb{E}\, D_n$ for the parameter values $\alpha = 0.1, \beta = 0.2$.
  \item[(d)] Verify your answer makes sense by checking the limiting cases $\alpha = \beta = 1/2$ (uniform iid) and $\alpha \to 0, \beta \to 0$ (nearly deterministic).
\end{enumerate}
\end{exercise}
