\documentclass[11pt]{article}
\usepackage[review]{acl}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\microtypesetup{expansion=false} % avoid pdftex scalable-font expansion error
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{url}

\title{Claudespeak: Characterizing the Stylistic Fingerprint of \\ a Frontier Language Model}

\author{Juan D.\,R. \\
  The University of Texas at Austin \\
  \texttt{juand-r@utexas.edu} \\}

\begin{document}
\maketitle

\begin{abstract}
Large language models are widely believed to have recognizable writing styles,
but most popular claims (e.g.\ that models overuse the word \emph{delve}) are
derived from \emph{ChatGPT} and applied indiscriminately to all models. We ask
what, specifically, characterizes the prose of one frontier model---Anthropic's
Claude Opus 4.8---relative to human writing and to seven other models, holding
topic constant via parallel corpora. Combining interpretable stylometry,
log-odds $n$-gram selection, a length-controlled regression, an interpretable
classifier, and suffix-automaton ($n$-gram novelty) mining, we find a stable
fingerprint that is \emph{structural, rhythmic, and interactional} rather than
lexical-in-the-stereotyped-sense. Claude is identifiable from prose alone
(classifier AUC $0.95$ vs.\ human/GPT-4o; $0.87$ vs.\ six modern models), driven
by markedly higher sentence-length variability, denser content-word usage, a
strong em-dash habit, and---most diagnostically---a verbatim offer-to-continue
closing move (``would you like me to go deeper into\ldots'') that appears dozens
of times in Claude and \emph{zero} times across humans and all other models.
We further show that the stereotyped ``delve'' vocabulary is a GPT (and
other-model) trait that Claude \emph{avoids}: across seven models in an essay
genre, Claude is the only model that never writes \emph{delve} and uses
\emph{crucial} least. We release the fully provenanced corpus and pipeline.
Sections on the effect of reasoning effort, Claude versions, and multi-turn
dynamics are forthcoming.
\end{abstract}

\section{Introduction}
The claim that text ``sounds like AI'' is now commonplace, and a folk
vocabulary of tells---\emph{delve}, \emph{tapestry}, \emph{crucial},
\emph{intricate}---circulates widely. Two problems undercut these claims as
characterizations of any \emph{particular} model. First, the evidence base is
overwhelmingly about ChatGPT \citep{kobak2024delving}, yet the conclusions are
applied to ``LLMs'' generically. Second, most analyses do not control topic: a
classifier that distinguishes model from human text may be learning the prompt
distribution, not the writing style.

We study a single frontier model, Anthropic's Claude Opus 4.8, and ask a
deliberately narrow question: \emph{holding the prompt fixed, what makes Claude's
prose identifiable relative to humans and to other models?} We build two
topic-matched parallel corpora in which the same prompts are answered by Claude,
by humans, and by up to six other models, and we attack the question with five
complementary methods. Our contributions:
\begin{enumerate}
  \itemsep0em
  \item A parallel-corpus protocol and a fully provenanced dataset
    (every generation stored with model, decoding configuration, token usage,
    and timestamp) that supports reuse without regeneration.
  \item Evidence that Claude's fingerprint is \emph{structural and rhythmic}
    (sentence-length burstiness, content-word density, em-dashes) and survives
    both markdown stripping and length control.
  \item Identification of a verbatim \emph{interactional} signature---an
    offer-to-continue closer---via log-odds and suffix-automaton mining, present
    in Claude and absent from all other sources.
  \item A reconciliation of the ``delve'' folklore: the stereotyped vocabulary
    is a GPT/other-model trait; Claude is the cleanest frontier model on it.
  \item A \emph{coverage taxonomy} (intent $\times$ interaction structure $\times$
    stance) and a reproducible audit showing that standard corpora occupy a
    narrow slice (single-turn, cooperative, informational); we argue that
    extending model characterization into the under-observed cells---multi-turn
    corrective, contested-stance, sensitive, emotional---is itself a
    contribution, and we release the audit tooling.
\end{enumerate}

\section{Related Work}
\paragraph{LLM stylometry and detection.} Distinguishing machine- from
human-written text is an active area, with large shared benchmarks
\citep{dugan2024raid} and classical authorship features such as function-word
distributions and Burrows's Delta \citep{burrows2002delta}. We use interpretable
features in this tradition, but our goal is \emph{characterization}---naming what
differs---rather than detection alone.

\paragraph{Excess vocabulary and ``delve.''} \citet{kobak2024delving} track
``excess words'' in 14M+ PubMed abstracts after ChatGPT's release, finding a
surge of style words (\emph{delves}, \emph{intricate}, \emph{crucial}). This
phenomenon is tied to ChatGPT specifically; popular accounts attribute the
``delve'' habit to OpenAI's RLHF annotation pipeline \citep{willison2024delve}.
A direct Claude-vs-GPT comparison reports the opposite lean for Claude---toward
casual, subjective phrasing \citep{personalization2025}. Our results corroborate
and quantify this on parallel data.

\paragraph{$n$-gram selection and novelty.} For surfacing characteristic
phrases we use the log-odds-ratio-with-informative-Dirichlet-prior estimator of
\citet{monroe2008fightin}, and for arbitrary-length span statistics we use
suffix-automaton tooling \citep{liu2024rustydawg} in the spirit of unbounded
$n$-gram modeling \citep{liu2024infinigram}.

\paragraph{Data.} Our human anchor is HC3 \citep{guo2023hc3}; our modern
multi-model contrast reuses published completions from AlpacaEval
\citep{alpacaeval2023}.

\section{Corpus}
We construct two parallel tracks (Table~\ref{tab:corpus}). In each, a fixed set
of prompts is answered by multiple sources, so that any feature separating Claude
reflects \emph{how} it writes rather than \emph{what} it was asked. Claude
(and GPT-4o on the HC3 track) are generated by us with adaptive thinking at
\texttt{effort=high}; all other cells are reused from the source datasets. Every
record stores full provenance---model and version, decoding configuration,
token usage, timestamp, and code commit---plus the verbatim API response.

\begin{table}[t]
\centering\small
\begin{tabular}{@{}lll@{}}
\toprule
Track & Prompts & Sources (same prompt each) \\
\midrule
HC3 & 200 & human, ChatGPT (GPT-3.5 era), \\
(human- & (5 dom.) & GPT-4o$^\dagger$, \textbf{Claude Opus 4.8}$^\dagger$ \\
anchored) & & \\[2pt]
Alpaca & 200 & gpt-4-turbo, gpt-4o, gemini-pro, \\
Eval & (instr.) & deepseek-67b, Qwen2-72B, \\
(modern) & & Llama-3-70B, \textbf{Claude Opus 4.8}$^\dagger$ \\
\bottomrule
\end{tabular}
\caption{The two parallel corpus tracks. $\dagger$ = self-generated; all other
cells reused. HC3 carries a human anchor; AlpacaEval carries the modern
multi-model contrast (no human, by design).}
\label{tab:corpus}
\end{table}

\subsection{Corpus statistics}
\label{sec:stats}
Table~\ref{tab:stats} summarizes volume and length. Responses are model-typical in
length (Claude median 230--267 words); the human HC3 answers are much shorter and
more variable (median 77, mean 120), which we account for via length controls
(\S\ref{sec:length}). Prompts are short (median 10--18 words) and roughly half are
phrased as questions. Per-domain volume and full distributions are in
Appendix~\ref{app:stats}. Two corpus-level register signals already foreshadow the
results: Claude's responses are markedly more \emph{interrogative} at the sentence
level (17\% of sentences on HC3, 10\% on AlpacaEval, vs.\ $\le$4\% for every other
source---the offer-closer, \S\ref{sec:ngram}), and least readable (Flesch $\approx$40
vs.\ 46--52), consistent with higher content-word density.

\begin{table}[t]
\centering\small
\begin{tabular}{@{}lrrrr@{}}
\toprule
Track & Prompts & Prompt & Claude resp. & Total \\
 & & wds (md) & wds (md) & words \\
\midrule
HC3 & 200 & 10 & 267 & 156k \\
AlpacaEval & 200 & 18 & 230 & 354k \\
\bottomrule
\end{tabular}
\caption{Corpus volume and length (median word counts). Totals sum all sources in
the track. Full per-source distributions: Appendix~\ref{app:stats}.}
\label{tab:stats}
\end{table}

\subsection{Coverage and its limits}
\label{sec:coverage}
Because style is context-dependent, we audit \emph{what behavior the prompts
elicit}, labeling every prompt by intent, speech act, register, tone, and topic,
plus interaction flags (\texttt{src/analyze\_coverage.py}; Table~\ref{tab:coverage}).
The corpus occupies a narrow region of the behavior space along several axes:
\begin{itemize}
  \itemsep0em
  \item \textbf{Speech act / mood.} Prompts are 74\% (HC3) / 58\% (AlpacaEval)
    \emph{interrogative} and 2\% / 20\% imperative; only $\sim$22\% are
    \emph{declarative}. The corpus almost never presents Claude with an
    \emph{assertion} to agree with, qualify, or contest.
  \item \textbf{Interaction.} Only $\sim$2\% of prompts imply multi-turn
    interaction, and essentially none are adversarial or corrective.
  \item \textbf{Intent.} Code, creative, opinion, and emotional intents are rare
    or absent, especially in HC3 (Table~\ref{tab:coverage}).
  \item \textbf{Stance / register / tone.} Prompts are predominantly neutral in
    tone and information-seeking; subjective or contested-stance prompts are a
    small minority.
\end{itemize}
Consequently the fingerprint we report should be read as Claude's
\emph{single-turn informational} voice. Behaviors gated on other cells---e.g.\ the
concession move ``you're right to push back on that,'' which requires a
\emph{declarative} challenge across turns---are out of scope here and motivate the
corpus expansion (more prompt variety across topic, register, tone, speech act,
and interaction structure) that we describe as future work
(\S\ref{sec:secondary}, Appendix~\ref{app:stats}).

\begin{table}[t]
\centering\small
\begin{tabular}{@{}lrr@{}}
\toprule
Intent (primary) & HC3 & AlpacaEval \\
\midrule
explanation & 95 & 34 \\
factual QA & 55 & 28 \\
advice/recommendation & 46 & 31 \\
instruction task & 2 & 29 \\
creative writing & 0 & 23 \\
math/reasoning & 0 & 15 \\
opinion/subjective & 0 & 12 \\
code & 0 & 10 \\
other (classif., edit, roleplay, & 2 & 18 \\
\quad personal, $\ldots$) & & \\
\midrule
\emph{flag:} multi-turn implied & 5 & 3 \\
\emph{flag:} subjective & 13 & 35 \\
\bottomrule
\end{tabular}
\caption{Prompt intent distribution and interaction flags (counts of 200 per
track). The corpus is single-turn, cooperative, and information-oriented.}
\label{tab:coverage}
\end{table}

\section{Methods}
We compute interpretable features per response (lexical: type-token ratio,
hapax rate, function-word rate; punctuation: em-dash, colon, question,
exclamation, emoji; syntax: mean sentence length and \emph{burstiness}, the
ratio of the standard deviation to the mean of words-per-sentence;
markdown/structure; and rhetoric). Rates are per 100 or 1000 words so that text
length does not mechanically dominate. We summarize source differences with
Cohen's $d$ (Claude vs.\ a comparison group). To separate \emph{voice} from
\emph{formatting}, we additionally compute all prose features after stripping
markdown markup. We mine characteristic $n$-grams with the
\citet{monroe2008fightin} log-odds estimator; quantify length-independence with
OLS (\S\ref{sec:length}); measure separability with $\ell_2$-regularized logistic
regression under 5-fold cross-validation; and mine arbitrary-length spans and
$n$-gram novelty with a suffix automaton \citep{liu2024rustydawg}.

\section{Results}

\subsection{A large signal, dominated at first by formatting}
The strongest raw separators between Claude and the other HC3 sources are
markdown-structural: markdown headers ($d{=}2.9$), bold ($d{=}2.3$), and bullets
($d{=}1.8$), which the human and old-ChatGPT cells use essentially never. This
is real but is a \emph{formatting} habit; the decisive question is whether
anything survives removing it.

\subsection{The voice survives markdown stripping}
\label{sec:prose}
With all markup stripped, Claude still separates strongly on prose
(Table~\ref{tab:prose}): it interleaves short and long sentences far more than
any other source (burstiness), writes denser content-word prose (lower
function-word rate), and retains a marked em-dash habit. These are the
load-bearing stylistic findings.

\begin{table}[t]
\centering\small
\begin{tabular}{@{}lcc@{}}
\toprule
Prose feature (markdown stripped) & $|d|$ & Claude vs.\ others \\
\midrule
Sentence burstiness & 1.69 & 0.86 vs.\ 0.36--0.55 \\
Function-word rate (lower) & 1.47 & 0.33 vs.\ 0.37--0.44 \\
Em-dash rate & 0.87 & 0.88 vs.\ $\le$0.40 \\
Colon rate & 0.84 & 1.41 vs.\ 0.24--1.69 \\
Question rate & 0.82 & 0.61 vs.\ 0.01--0.28 \\
\bottomrule
\end{tabular}
\caption{HC3 track, Claude vs.\ pooled others, prose only. Mean $|$Cohen's $d|$
and group means.}
\label{tab:prose}
\end{table}

\subsection{The signature is not a length artifact}
\label{sec:length}
Claude writes longer answers (mean 253 words vs.\ 120 for humans, 228 for
GPT-4o), so we test whether the prose effects merely track length. Within every
word-count quartile the effects persist, and a length-controlled regression
($\text{feature}\sim\text{is\_claude}+z(\text{n\_words})$) leaves the Claude
coefficient large and highly significant: burstiness $t{=}20.6$, function-word
rate $t{=}{-}14.4$, em-dash $t{=}12.6$, question rate $t{=}4.6$. Type-token ratio
is length-sensitive and we treat lexical-diversity claims as length-conditioned.

\subsection{The interactional signature: an offer-to-continue closer}
\label{sec:ngram}
Data-driven $n$-gram mining \citep{monroe2008fightin} surfaces a clear
interactional fingerprint rather than a content-word list. The most Claude-like
trigrams (HC3) are \emph{would you like}, \emph{like me to}, \emph{me to
explain}, \emph{in more detail}, \emph{to go deeper}; the anti-signature---terms
Claude \emph{underuses}---is led by the formal frame \emph{it is important to}.
Suffix-automaton mining (\S\ref{sec:dawg}) confirms this at arbitrary length.

\subsection{The ``delve'' vocabulary is a GPT trait Claude avoids}
\label{sec:delve}
On the AlpacaEval track---an essay/instruction genre that \emph{does} elicit the
stereotyped words---Claude is the cleanest of seven models
(Table~\ref{tab:delve}). It is the only model that never writes \emph{delve} or
\emph{delves}, and it uses \emph{crucial} least of all (0.04 per 1k vs.\ 0.34--0.42
for the GPT models and Qwen). Claude's own lexical lean is conversational
(``actually'' occurs $\sim$6$\times$ more often in Claude than GPT-4o on HC3),
consistent with \citet{personalization2025}; this is a secondary signal beneath
the structural one.

\begin{table}[t]
\centering\small
\begin{tabular}{@{}lccccc@{}}
\toprule
per-1k & Claude & g4t & g4o & qwen & llama \\
\midrule
delve   & \textbf{0} & .018 & .018 & 0 & .017 \\
delves  & \textbf{0} & .037 & .089 & .019 & .034 \\
crucial & \textbf{.039} & .422 & .340 & .346 & .185 \\
intricate & \textbf{0} & 0 & 0 & .019 & .050 \\
pivotal & \textbf{0} & .073 & .018 & .038 & 0 \\
\bottomrule
\end{tabular}
\caption{Frequency (per 1000 words) of stereotyped ``excess'' words on the
AlpacaEval essay track. g4t = gpt-4-turbo, g4o = gpt-4o. Claude is at or near
the minimum for every term. (gemini/deepseek omitted for space; same pattern.)}
\label{tab:delve}
\end{table}

\subsection{Separability}
\label{sec:clf}
An interpretable classifier confirms a stable, learnable voice
(Table~\ref{tab:clf}). Stripping markdown reduces AUC only modestly, so the
classifier is not merely reading layout; and its top coefficients are exactly the
features of \S\ref{sec:prose} (burstiness, function-word density, em-dash,
question rate), an independent-method confirmation.

\begin{table}[t]
\centering\small
\begin{tabular}{@{}lcc@{}}
\toprule
Track & with formatting & prose only \\
\midrule
HC3 (vs.\ human/GPT-4o/old-ChatGPT) & 0.982 & 0.946 \\
AlpacaEval (vs.\ 6 modern models) & 0.901 & 0.867 \\
\bottomrule
\end{tabular}
\caption{Claude-vs-rest ROC-AUC, logistic regression, 5-fold CV.}
\label{tab:clf}
\end{table}

\subsection{Arbitrary-length spans and $n$-gram novelty}
\label{sec:dawg}
Suffix-automaton mining \citep{liu2024rustydawg} gives the most concrete form of
the signature. Spans up to \emph{eight words} occur dozens of times in Claude and
\emph{zero} times across humans plus all six other models combined:
\emph{would you like me to explain any} ($\times$20), \emph{would you like me to
go deeper into} ($\times$14), and on AlpacaEval \emph{would you like me to expand
on any} ($\times$7). Claude also has the highest $n$-gram novelty of any source
(e.g.\ bigram novelty 0.519 vs.\ 0.27--0.36 for the six AlpacaEval models): its
exact phrasing overlaps least with the pool. We read novelty as relative
distinctiveness, not absolute creativity (see Limitations).

\subsection{Secondary analyses (in progress)}
\label{sec:secondary}
Three planned analyses are not yet included; designs and placeholder tables
appear in Appendix~\ref{app:placeholders}.
\begin{itemize}
  \itemsep0em
  \item \textbf{Effect of reasoning effort/thinking.} Whether the fingerprint
    intensifies or attenuates across \texttt{effort}$\in\{$low, medium, high$\}$
    and with thinking disabled. \emph{[To be completed; Appendix~\ref{app:effort}.]}
  \item \textbf{Across Claude versions.} Intra-Claude drift across model versions
    (e.g.\ Opus/Sonnet/Haiku and older releases). \emph{[To be completed;
    Appendix~\ref{app:versions}.]}
  \item \textbf{Multi-turn and self-interaction.} Turn-dependent traits and the
    Claude-to-Claude ``spiritual bliss'' attractor reported in
    \citep{anthropic2025claude4}. \emph{[To be completed;
    Appendix~\ref{app:multiturn}.]}
\end{itemize}

\section{Discussion}
Across two corpora, four methods, and eight comparison sources, the same
fingerprint recurs and survives every control: Claudespeak is
markdown-scaffolded, rhythmically varied, content-dense, and
engagement-and-offer-oriented prose---and specifically \emph{not} the ``delve''
register. A methodological lesson follows: borrowed AI-word lists largely measure
\emph{GPT}, and characterizing a specific model requires data-driven discovery
against topic-matched references.

\section*{Limitations}
We study a single Claude configuration (Opus 4.8, \texttt{effort=high}); the
effort, version, and multi-turn axes are not yet measured (\S\ref{sec:secondary}).
Crucially, our coverage audit (\S\ref{sec:coverage}) shows the corpus is
single-turn, cooperative, and information-oriented, so the reported fingerprint is
Claude's \emph{single-turn informational} voice; context-gated behaviors
(corrective concession, refusal, emotional support, code/creative registers) are
out of scope and are the target of the planned expansion (English only).
Reused contrast completions are
their published vintages (e.g.\ gemini-pro~1.0, Qwen2-72B, GPT-4-turbo/4o~2024),
modern-ish but not the absolute latest. $n$-gram novelty is measured relative to
our corpus pool, not a large external reference, and may be inflated for Claude
as the lone out-of-cluster (and only freshly generated) source. The HC3 human
anchor is informal web text, so part of the human--Claude gap reflects register
rather than authorship.

\section*{Ethics Statement}
This work characterizes model writing style using public datasets and model
outputs; it involves no human subjects or private data. Stylistic fingerprints
can aid provenance and detection but could also inform evasion; we report
aggregate, interpretable features rather than a deployable detector.

\bibliography{custom}

\appendix

\section{Feature definitions}
\label{app:features}
Full definitions of all interpretable features (rates, burstiness, markdown and
rhetorical markers) are provided with the released code (\texttt{src/features.py}).

\section{Full corpus statistics}
\label{app:stats}
All statistics here are reproducible via \texttt{src/dataset\_stats.py} (volume,
lengths, sentence mood, register) and \texttt{src/analyze\_coverage.py} (intent).

\paragraph{Response length by source (words).}
HC3: human 120/77 (mean/median), ChatGPT 176/178, GPT-4o 228/239, Claude
254/267. AlpacaEval: deepseek-67b 186/184, gemini-pro 222/234, Claude 254/230,
Qwen2-72B 260/259, gpt-4-turbo 272/285, gpt-4o 280/285, Llama-3-70B 297/308.

\paragraph{Volume by domain (prompts; total response words, all sources).}
HC3: general 80/47k, finance 40/40k, cs\_ai 40/35k, medicine 40/33k. AlpacaEval:
selfinstruct 64/79k, oasst 49/96k, helpful\_base 35/70k, koala 35/67k, vicuna
17/43k.

\paragraph{Sentence mood (\% of response sentences).} Interrogative sentences:
Claude 17\% (HC3) / 10\% (AlpacaEval) vs.\ 1--4\% for all other sources;
declarative 75\% (Claude, HC3) vs.\ 95--97\% for others. Imperative and
exclamatory are $\le$2\% and $\le$7\% respectively.

\paragraph{Register.} First-/second-person pronoun rates per 100 words and Flesch
Reading Ease are in \texttt{stats\_register.csv}; Claude shows the lowest Flesch
(\,$\approx$40, denser/harder) of all sources on both tracks.

\paragraph{Prompt variety (audit, counts of 200 per track).} The prompts are
narrow on every axis (Table~\ref{tab:variety}): almost all are
\emph{questions} or \emph{directives} (assertions $\le$1); tone is
overwhelmingly \emph{neutral} (82--87\%); register is neutral/casual with almost
no \emph{formal} prompts; and topics concentrate (HC3: health 28\%, science 24\%,
finance 20\%). This uniformity---not insufficient sample size---is the coverage
problem the expansion targets.

\begin{table}[t]
\centering\small
\begin{tabular}{@{}llrr@{}}
\toprule
Axis & Value & HC3 & Alpaca \\
\midrule
speech act & question & 150 & 106 \\
 & directive & 46 & 93 \\
 & assertion & 1 & 0 \\
register & neutral & 130 & 130 \\
 & casual & 70 & 62 \\
 & formal & 0 & 7 \\
tone & neutral & 165 & 175 \\
 & emotional & 20 & 5 \\
 & playful / urgent & 1/14 & 18/1 \\
\bottomrule
\end{tabular}
\caption{Prompt variety along speech act, register, and tone (counts of 200).
Assertions, formal register, and non-neutral tones are nearly absent. Topic and
intent breakdowns: \texttt{coverage\_labels.csv}.}
\label{tab:variety}
\end{table}

\paragraph{Coverage expansion (planned).} To move beyond the single-turn
informational slice (\S\ref{sec:coverage}) we will (i) sample real first-turn
prompts from WildChat / LMSYS-Chat-1M to populate under-covered intents
(code, creative, opinion, emotional); (ii) \emph{construct} multi-turn corrective
scenarios (user challenges or corrects Claude) to elicit concession / push-back
behavior; and (iii) add a small, defensively-framed sensitive/refusal set.
Targets are $\sim$100--150 prompts per intent cell and $\sim$80--120 corrective
scenarios; the audit tooling re-runs on the expanded corpus.

\section{Placeholders for secondary analyses}
\label{app:placeholders}

\subsection{Effect of reasoning effort}
\label{app:effort}
\emph{Planned design.} Regenerate the 200 HC3 prompts (and 200 AlpacaEval
prompts) under \texttt{effort}$\in\{$low, medium, high$\}$ and with thinking
disabled, holding all else fixed; recompute the \S\ref{sec:prose} features and
the \S\ref{sec:clf} classifier within each setting; report whether burstiness,
density, em-dash, and the offer-closer scale with effort.
\emph{[Results table to be inserted.]}

\subsection{Across Claude versions}
\label{app:versions}
\emph{Planned design.} Generate the same prompts with additional Claude versions
(e.g.\ Sonnet/Haiku, and an older 3.x), then measure intra-Claude drift in the
fingerprint features and the offer-closer rate.
\emph{[Results table to be inserted.]}

\subsection{Multi-turn and self-interaction}
\label{app:multiturn}
\emph{Planned design.} (i) Two-turn parallel probes (e.g.\ MT-Bench-style) to
measure turn-dependent traits comparatively; (ii) elicited Claude-to-Claude
conversations to test the ``spiritual bliss'' attractor
\citep{anthropic2025claude4} quantitatively (e.g.\ trajectory of
consciousness/gratitude vocabulary and vocabulary collapse over turns).
\emph{[Results table to be inserted.]}

\end{document}
